---
title: "Homework 03 - Functions, Profiling, and Rcpp"
format:
  html:
    embed-resources: true
---

# Due date

Thursday, October 31

# Background

For this assignment, you'll be quested with speeding up some code using what
you have learned about vectorization and Rcpp.

## Part 1: Vectorizing code

The following functions can be written to be more efficient without using
parallel computing:

1. This function generates a `n x k` dataset with all its entries distributed
Poisson with mean `lambda`.

```{r}
#| label: p1-fun1
#| eval: false
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n)
    x <- rbind(x, rpois(k, lambda))
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  # YOUR CODE HERE
}

# Benchmarking
bench::mark(
  fun1(),
  fun1alt(), relative = TRUE, check = FALSE
)
```

2. Like before, speed up the following functions (it is OK to use StackOverflow)

```{r}
#| label: p1-fun2
#| eval: false
# Total row sums
fun1 <- function(mat) {
  n <- nrow(mat)
  ans <- double(n) 
  for (i in 1:n) {
    ans[i] <- sum(mat[i, ])
  }
  ans
}

fun1alt <- function(mat) {
  # YOUR CODE HERE
}

# Cumulative sum by row
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}

fun2alt <- function(mat) {
  # YOUR CODE HERE
}

# Use the data with this code
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)

# Test for the first
bench::mark(
  fun1(dat),
  fun1alt(dat), relative = TRUE
)

# Test for the second
bench::mark(
  fun2(dat),
  fun2alt(dat), relative = TRUE
)
```

3.  Find the column max (hint: Check out the function `max.col()`).

```{r}
#| label: p1-fun3
#| eval: false
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

fun2alt <- function(x) {
  # YOUR CODE HERE
}

# Benchmarking
bench::mark(
  fun2(),
  fun2alt(), relative = TRUE
)
```

## Part 2: C++

As we saw in the Rcpp week, vectorization may not be the best solution. For this
part, you must write a function using Rcpp that implements the [scale-free algorithm](https://gvegayon.github.io/networks-udd2024/02-random-graphs.html#scale-free-networks). The following code implements this in R:

```{r}
## Model parameters
n <- 500
m <- 2

## Generating the graph
set.seed(3312)
g <- matrix(0, nrow = n, ncol = n)
g[1:m, 1:m] <- 1
diag(g) <- 0

## Adding nodes
for (i in (m + 1):n) {

  # Selecting the nodes to connect to
  ids <- sample(
    x       = 1:(i-1), # Up to i-1
    size    = m,       # m nodes
    replace = FALSE,   # No replacement
    # Probability proportional to the degree
    prob    = colSums(g[, 1:(i-1), drop = FALSE])
    )

  # Adding the edges
  g[i, ids] <- 1
  g[ids, i] <- 1

}

## Visualizing the degree distribution
library(ggplot2)
data.frame(degree = colSums(g)) |>
  ggplot(aes(degree)) +
  geom_histogram() +
  scale_x_log10() +
  labs(
    x = "Degree\n(log10 scale)",
    y = "Count"
  )
```

Rewrite the function that generates the scale-free network using Rcpp.