---
format: gfm
---

![GitHub last commit](https://img.shields.io/github/last-commit/UofUEpiBio/PHS-7045-egga)

# Latest papers on Agent-Based Modeling in PubMed

This repository is an *Example using GitHub Actions* \[egge\] for the Advanced Programming with R class. The project consists of a quarto document updated once a week or whenever there's a push or pull request. It performs a search on PubMed looking for the [latest papers about Agent-Based Models](https://pubmed.ncbi.nlm.nih.gov/?term=agent-based+model&sort=date){target="_blank"}. The last update was on `r Sys.time()`.

```{r}
#| label: data-extract
#| echo: false
library(xml2)

doc <- read_html("https://pubmed.ncbi.nlm.nih.gov/?term=agent-based+model&sort=date")

# Titles
titles <- xml_find_all(doc, xpath = '//*[@class="docsum-title"]')
titles <- xml_text(titles)
titles <- trimws(titles, which = "both")

# Authors
authors <- xml_find_all(doc, xpath = '//*[@class="docsum-authors full-authors"]')
authors <- xml_text(authors)

# Citation
citation <- xml_find_all(doc, xpath = '//*[contains(@class, "full-journal-citation")]')
citation <- xml_text(citation)

# Abstract 
abstract <- xml_find_all(doc, xpath = '//*[@class="full-view-snippet"]')
abstract <- xml_text(abstract)
abstract <- trimws(abstract, which = "both") #Removing trailing and leading space
```


```{r}
#| results: asis
#| label: printing
#| echo: false
for (i in 1:length(titles)) {
  
  cat(sprintf("%i. **%s**\n\n    %s\n\n    %s\n\n    >%s\n\n", i, titles[i], authors[i], citation[i], abstract[i]))
  
}
```


