---
title: Lab 05 - C++
author: George G. Vega Yon, Ph.D.
date-modified: 2024-09-16
date: 2024-09-19
format:
  html:
    embed-resources: false
---

## Learning goals

- Practice class programming in C++.
- Be able to map R functions to C++.
- Practice your GitHub skills.

## Lab description

For this lab, we will implement a class representing a binomial distribution:

$$
\Pr{\left(Y=k; n, p\right)} = {n \choose  k}p^k(1-p)^{n-k}
$$

Implement the following class representing a binomial distribution:

```cpp
class Binom {
private:
  int n;
  double p;
  
public:
  // Binom(int n_, double p_) : n(n_), p(p_) {};
  Binom(int n_, double p_) {
    n = n_;
    p = p_;
  };
  int factorial(int k) const;
  double choose(int a, int b) const;
  double dbinom(int k) const;
  void print(int k) const;
};
```

Remember that to implement the function, you have to use the following pattern:

```cpp
inline [return type] [class name]::[function name]([arguments]) {
    // Your code here
}
```

### Where to look things up

C++ has no equivalent of R's `?pow` or `apropos()`. The references below can be used to look up functions, their signatures, and the headers they live in.

- [**cppreference.com**](https://en.cppreference.com/w/cpp) is the reference
  the C++ community actually uses. Two pages will cover most of this lab:
  [Standard library headers](https://en.cppreference.com/w/cpp/header), which
  lists what lives where, and [Common mathematical
  functions](https://en.cppreference.com/w/cpp/numeric/math), which documents
  everything in `<cmath>` (`pow`, `sqrt`, `exp`, `log`, ...). The pages open with the full set of
  overloads and can look intimidating. Scroll down to *Parameters*, *Return
  value*, and especially the compilable *Example* at the bottom.
- [**learncpp.com**](https://www.learncpp.com/) is a free, well-sequenced
  tutorial. Use it when you want to understand a concept rather than look up a
  signature.

### Exercise 1: Factorial

Write a program that implements the `factorial` function. It should match the following results in R

```{r}
factorial(1:10)
```

You can use the following code to test your function:

```cpp
Binom b(10, 0.5);
for (int i = 0; i < 10; i++) {
    std::cout << b.factorial(i) << std::endl;
}
```

### Exercise 2: Choose

Write a program that implements the `choose` function. It should match the following results in R

```{r}
choose(10, 1:10)
```

You can use the following code to test your function:

```cpp
Binom b(10, 0.5);
for (int i = 0; i < 10; i++) {
    std::cout << b.choose(10, i) << std::endl;

}
```

::: {.callout-tip collapse="true"}
## Tip (only check this if your output looks wrong)

If your results come out as `0`, or as whole numbers where R gives you
something else, the likely cause is **integer division**. In C++ the `/`
operator looks at the *types* of its operands, not at what you meant. If both
sides are `int`, you get integer division and the remainder is thrown away:

```cpp
int a = 7, b = 2;
a / b;        // 3, not 3.5 -- the .5 is discarded
```

Since `factorial` returns an `int`, an expression such as

```cpp
return factorial(a) / (factorial(b) * factorial(a - b));
```

divides one `int` by another, truncates, and only then converts the result to
the `double` the function returns. The fix is to make at least one operand a
`double` *before* the division happens, which is what `static_cast` is for:

```cpp
return static_cast<double>(factorial(a)) /
  (static_cast<double>(factorial(b)) * static_cast<double>(factorial(a - b)));
```

`static_cast<T>(expression)` converts `expression` to type `T` at compile time.
It is the C++ way of writing the C-style cast `(double) x`, and it is preferred
because it is easy to search for and because the compiler checks that the
conversion actually makes sense. Note that the cast has to be on the operands,
not on the result: `static_cast<double>(a / b)` is too late -- the truncation
has already happened inside the parentheses.

One more thing to watch: think about what your function should return when
`b < 0` or `b > a`. R's `choose` gives `0` there.
:::

### Exercise 3: implement the binom

Write a program that implements the `dbinom` function. It should match the following results in R

```{r}
dbinom(0:10, 10, 0.5)
```

You can use the following code to test your function:

```cpp
Binom b(10, 0.5);
for (int i = 0; i < 10; i++) {
    std::cout << b.dbinom(i) << std::endl;
}
```

::: {.callout-tip collapse="true"}
## How do I raise a number to a power?

C++ has no `^` operator for exponentiation -- `p ^ k` will not do what you want
(`^` is bitwise XOR, and it does not even compile for `double`). Use `std::pow`
from the `<cmath>` header instead:

```cpp
#include <cmath>

std::pow(0.5, 3);   // 0.125
std::pow(2.0, 10);  // 1024
```

`std::pow(base, exponent)` returns a `double`, and both arguments are converted
to floating point first, so passing an `int` exponent such as your `k` is fine.

For the binomial density you need two of these, one for the successes and one
for the failures:

$$
P(Y = k) = \binom{n}{k} \, p^k \, (1 - p)^{\,n - k}
$$

so the second term has exponent `n - k`. Both are `int`s, which is no problem:
the subtraction is done in integer arithmetic and the result is then converted
for `std::pow`.

Two small things to watch:

- `std::pow` lives in `<cmath>`, so make sure that header is included at the
  top of your file.
- Mind your parentheses. `std::pow(1 - p, n - k)` is the failure term;
  something like `std::pow(1 - p, n) - k` compiles just as happily and is not
  what you meant, since everything involved is a number.
:::

### Exercise 4: Print

Write a program that implements the `print` function. It should match the following results in R

```{r}
sprintf(
  "P(Y=%-2d; n=%d, p=%.2f) = %.4f",
  0:10, 10, 0.5, dbinom(0:10, 10, 0.5)
  ) |>
  cat(sep = "\n")
```