---
title: "How R Functions Really Work"
subtitle: "PHS 7045: Advanced Programming — Fall 2026"
format:
  revealjs:
    embed-resources: true
    slide-number: true
    scrollable: true
author: "PHS 7045 Teaching Team"
---

```{r}
#| echo: false
#| label: setup
#| include: false
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, results = "hold",
                      fig.width = 7, fig.height = 5)
```

## Today's goals

::: {.r-fit-text}

You already know how to *write* a function. Today is about **why R's functions
behave differently from every other language you know** — the things that are
particular about R.

1. **Everything is a function call** — the two slogans of R.
2. **Functions are objects** — first-class, anonymous, returned, stored.
3. **Lazy evaluation** — arguments are *promises*, not values.
4. **Scoping & closures** — functions carry their environment; environments are mutable.
5. **Functionals** — the functional-programming payoff (and the bridge to parallel computing).
6. **Non-standard evaluation** — how `dt[, temp]` and `filter(df, x > 1)` actually work.

**The spine:** lazy evaluation is the hinge. Everything before it explains *how*
arguments and scopes behave; NSE is what you get when you capture a promise
instead of forcing it.

::: {.fragment .fade-up}
*Assumed:* the R essentials pre-read (vectors, control flow, writing a basic function).
:::

:::

# Functional programming, in context

## Four ways to organize a program

::: {.r-fit-text}

A programming *paradigm* is a way of thinking about what a program **is**. They
differ in what they put at the center:

| Paradigm | Organized around | The program is… | Examples |
|---|---|---|---|
| **Imperative / procedural** | statements that change state | a sequence of steps: *how* to do it | C, Python, R |
| **Object-oriented** | objects bundling data + methods | objects exchanging messages, mutating state | Java, C++, Python |
| **Functional** | functions and their composition | expressions to evaluate, avoiding mutable state | Haskell, Lisp, R |
| **Logic / declarative** | facts, rules, and goals | *what* is true; the engine finds *how* | Prolog, SQL |

::: {.fragment .fade-up}
These are **lenses, not boxes**. Most real languages — R included — mix several.
The question is which lens the language makes *natural*.
:::

:::

## Functional programming — and where R sits

::: {.r-fit-text}

Functional programming rests on a few ideas:

- **Functions are first-class** — store them, pass them, return them.
- **Composition over iteration** — build behavior by combining small functions
  (`Map`, `Reduce`, `purrr::map`) instead of hand-written loops.
- **Avoid side effects** — prefer returning new values to mutating in place
  (recall *copy-on-modify*).

The same task, two mindsets — sum the squares of `1:n`:

```r
# Imperative: describe every step, mutate an accumulator
total <- 0
for (x in 1:n) total <- total + x^2

# Functional: compose functions, no mutable state
sum(vapply(1:n, \(x) x^2, numeric(1)))
```

::: {.fragment .fade-up}
**Where R sits:** R is *multi-paradigm* — it has three OOP systems (S3, S4, R6)
and you will write plenty of loops. But at its **core R is functional**: `+` is a
function, functions are values, and the vectorized `*apply` idioms are functional
programming. That is the side of R this week is about.
:::

:::

# 1. Everything is a function call

## The two slogans of R

::: {.r-fit-text}

> To understand computations in R, two slogans help:
>
> - Everything that **exists** is an object.
> - Everything that **happens** is a function call.
>
> — John Chambers

Operators, `if`, subsetting, even assignment are *function calls in disguise*.

```{r}
#| label: fn-call-1
`+`(2, 3)          # `+` is an ordinary function
`if`(2 > 1, "yes", "no")   # if returns a value
`[`(c(10, 20, 30), 2)      # subsetting is a function too
```

:::

---

::: {.r-fit-text}

Because operators are just functions, **you can define your own**:

```{r}
#| label: fn-call-2
`%+%` <- function(a, b) paste0(a, b)
"data" %+% ".table"
```

::: {.fragment .fade-up}
**Why this matters:** if operators and `if` are functions, then functions are
nothing special — they are values you can store, pass, and return. That is the
foundation for everything else today.
:::

:::

# 2. Functions are objects

## First-class functions

::: {.r-fit-text}

A function is a value like any other. You can put it in a list, pass it as an
argument, or return it from another function.

```{r}
#| label: first-class-1
funs <- list(
  root   = sqrt,
  square = function(x) x^2,
  neg    = function(x) -x
)
funs$square(5)
```

```{r}
#| label: first-class-2
# A function that takes a function and returns a function
apply_twice <- function(f) function(x) f(f(x))
add1 <- function(x) x + 1
apply_twice(add1)(10)
```

:::

---

::: {.r-fit-text}

**Anonymous functions** have no name. Since R 4.1 there is a shorthand, `\(x)`:

```{r}
#| label: anon
(\(x) x^2)(4)          # same as (function(x) x^2)(4)
```

::: {.fragment .fade-up}
We will lean on anonymous functions constantly once we reach *functionals*.
:::

:::

# 3. Lazy evaluation

## Arguments are promises, not values

::: {.r-fit-text}

In most languages, every argument is computed **before** the call runs (*eager*).
R is different: an argument is a **promise** — an unevaluated expression plus the
environment to evaluate it in. It is forced **only when first used**.

```{r}
#| label: lazy-1
f <- function(x, y) x     # note: y is never used
f(1, stop("boom"))        # no error — y is never forced!
```

::: {.fragment .fade-up}
In Python or C, `stop("boom")` would run first and blow up. In R the promise for
`y` is simply never asked for.
:::

:::

## Question 1: What does this print? {.smaller}

::: {.r-fit-text}

```r
g <- function(x, y) {
  x * 2
}
g(5, stop("this never runs"))
```

::: {.fragment .fade-up}
`10` — the body only touches `x`. `y`'s promise is never forced, so `stop()`
never runs.
:::

:::

## Consequence: defaults can see other arguments

::: {.r-fit-text}

Because a default is a lazy promise evaluated *inside* the function, it can
reference other arguments — something no eager language allows.

```{r}
#| label: lazy-default
summarise_vec <- function(x, n = length(x)) {
  sprintf("mean of %d values: %.2f", n, mean(x))
}
summarise_vec(c(2, 4, 6))    # n defaults to length(x)
```

:::

## The classic gotcha: closures in a loop

::: {.r-fit-text}

Each function below captures the *promise* for `i`, which is forced later — after
the loop has already finished.

```{r}
#| label: lazy-gotcha
make_fns <- function() {
  fns <- vector("list", 3)
  for (i in 1:3) fns[[i]] <- function() i
  fns
}
fns <- make_fns()
c(fns[[1]](), fns[[2]](), fns[[3]]())    # 3 3 3, not 1 2 3!
```

::: {.fragment .fade-up}
`force(i)` evaluates the promise **now**. Functionals like `lapply()` give each
iteration its own `i`, so they sidestep the bug entirely:

```{r}
#| label: lazy-fix
fns2 <- lapply(1:3, function(i) function() i)
c(fns2[[1]](), fns2[[2]](), fns2[[3]]())  # 1 2 3
```
:::

:::

# 4. Scoping & closures

## Lexical scoping: functions carry their environment

::: {.r-fit-text}

A function remembers **where it was defined**, not where it is called. A function
plus the environment it captured is called a **closure**.

```{r}
#| label: closure-factory
# A "function factory": a function that manufactures functions
power <- function(exp) {
  function(x) x^exp
}
square <- power(2)
cube   <- power(3)
c(square(4), cube(2))
```

::: {.fragment .fade-up}
`square` and `cube` are the *same* body, but each closed over a different `exp`.
:::

:::

## Environments are mutable (reference semantics)

::: {.r-fit-text}

Vectors and lists are **copy-on-modify** (recall module 04). Environments are
the exception: they are modified **in place**.

```{r}
#| label: env-counter
# `<<-` assigns in the enclosing environment, so state persists between calls
new_counter <- function() {
  n <- 0
  function() {
    n <<- n + 1
    n
  }
}
count <- new_counter()
c(count(), count(), count())    # 1 2 3 — the closure keeps state
```

```{r}
#| label: env-ref
e <- new.env(); e$value <- 1
modify <- function(env) env$value <- 99   # no copy is made
modify(e)
e$value                                    # 99 — modified in place
```

:::

## Aside: replacement functions

::: {.r-fit-text}

Another R oddity that ties back to copy-on-modify: assigning into a function call
is secretly a function call.

```{r}
#| label: replacement
x <- c(1, 2, 3)
names(x) <- c("a", "b", "c")   # secretly calls `names<-`(x, value)
x
```

`names(x) <- v` is really `x <- \`names<-\`(x, v)` — which is why it produces a
*modified copy* rather than mutating in place.

:::

# 5. Functionals

## The functional-programming payoff

::: {.r-fit-text}

A **functional** takes a function as input and applies it over data — replacing
explicit loops with a single, declarative call.

```{r}
#| label: func-lapply
lapply(1:3, function(x) x^2)      # apply over a list, get a list back
```

```{r}
#| label: func-vapply
vapply(1:3, function(x) x^2, numeric(1))   # type-safe: you declare the shape
```

::: {.fragment .fade-up}
Prefer `vapply()` over `sapply()` in code you keep — it fails loudly instead of
silently returning the wrong type.
:::

:::

## More functionals: Map, Reduce, Filter

::: {.r-fit-text}

```{r}
#| label: func-map
Map(function(x, y) x + y, 1:3, 10:12)   # iterate over several inputs at once
```

```{r}
#| label: func-reduce
Reduce(`+`, 1:5)                        # fold a binary op across a vector -> 15
Reduce(`+`, 1:5, accumulate = TRUE)     # keep the running total
```

```{r}
#| label: func-filter
Filter(function(x) x %% 2 == 0, 1:10)   # keep elements passing a predicate
```

The tidyverse offers a type-stable family via **purrr**:

```r
library(purrr)
map_dbl(1:3, \(x) x^2)        # 1 4 9  (like vapply, always a double vector)
map2_dbl(1:3, 10:12, `+`)     # 11 13 15
```

:::

## Why we care: functionals are the bridge to parallelism

::: {.r-fit-text}

Once your iteration is a functional, going parallel is a **one-line change** —
because a function is just an object you can hand to a different engine.

```r
# today
lapply(files, process)

# in a few weeks — same shape, now parallel / on the cluster
parallel::mclapply(files, process, mc.cores = 4)
future.apply::future_lapply(files, process)
```

::: {.fragment .fade-up}
This only works because functions are first-class objects. Everything from slide
1 has been building to this.
:::

:::

# 6. Non-standard evaluation

## Capturing code instead of values

::: {.r-fit-text}

Lazy evaluation, taken one step further: instead of **forcing** a promise to get
its value, a function can inspect the **expression** itself with `substitute()`.

```{r}
#| label: nse-1
g <- function(x) deparse(substitute(x))
g(a + b * 2)     # "a + b * 2" — a and b need not even exist
```

```{r}
#| label: nse-standard-vs-not
h_standard <- function(x) x              # sees the VALUE
h_nse      <- function(x) substitute(x)  # sees the CODE
h_nse(1 + 1)                             # `1 + 1`, unevaluated
```

This is **non-standard evaluation (NSE)**: the function reads the code you typed,
not the value it produces.

:::

## This is the "magic" you have already been using

::: {.r-fit-text}

Every place you type a **bare, unquoted name** is NSE at work:

```r
subset(df, temp > 20)            # `temp` is not a variable in your workspace...
dplyr::filter(df, temp > 20)     # ...it is a bare name, captured then evaluated
dt[, .(temp, wind)]              # data.table does exactly this
lm(y ~ x, data = df)             # y ~ x is a captured expression (a formula)
library(dplyr)                   # bare name, not the string "dplyr"
```

None of these could work under standard evaluation — `temp` would be looked up
in the global environment and throw *"object not found."*

:::

## Build your own `subset()` in two lines

::: {.r-fit-text}

The whole trick: **capture** the expression, then **evaluate it in a chosen
environment** (the data frame).

```{r}
#| label: nse-subset
my_subset <- function(data, condition) {
  cond <- substitute(condition)          # 1. capture the code, unevaluated
  rows <- eval(cond, envir = data)       # 2. evaluate it INSIDE `data`
  data[rows, , drop = FALSE]
}
head(my_subset(mtcars, mpg > 30))        # `mpg` resolves to a column, not a global
```

:::

## The catch: NSE is a footgun when you program over it

::: {.r-fit-text}

Great interactively, painful programmatically. Watch this **silently wrong**
result:

```{r}
#| label: nse-catch
my_var <- "mpg"
nrow(my_subset(mtcars, my_var > 30))   # 32 — ALL rows!
nrow(mtcars)
```

`my_var` was captured as *code*, not the string it holds, so the condition
compared the literal name — a "confident, plausible, wrong" answer.

::: {.fragment .fade-up}
Escaping this — telling the function "evaluate this one normally" — is exactly
what tidyverse's `{{ }}` / `.data[[ ]]` and data.table's `env=` exist to solve.
:::

:::

# Wrap-up

## The arc of the week

::: {.r-fit-text}

1. **Everything is a function call** → functions are not special.
2. **Functions are objects** → store them, pass them, return them.
3. **Lazy evaluation** → arguments are promises, forced on first use.
4. **Closures** → functions carry (and can mutate) their environment.
5. **Functionals** → apply functions over data; the bridge to parallelism.
6. **NSE** → force the promise's *expression* instead of its *value*.

::: {.fragment .fade-up}
**One sentence:** lazy evaluation is the hinge — force a promise and you get a
value; grab its expression and you get NSE.
:::

:::

## Where this goes next

::: {.r-fit-text}

- **Parallel computing:** your `lapply()` becomes `mclapply()` / `future_lapply()`.
- **`data.table` & `dplyr`:** now you know *why* bare column names work.
- **R packages:** closures, `...`, and functionals are the building blocks of a
  clean API.

**Lab:** build a *function factory*, drive it with a *functional*, and write one
small *NSE helper* (a bare-column `summarise`, or a `deparse(substitute())`
error-message helper).

:::

## Session info {.smaller}

```{r}
#| label: sessioninfo
sessionInfo()
```
