---
title: "R Essentials and Functional Programming"
subtitle: "PHS 7045: Advanced Programming — Fall 2026"
format:
  revealjs:
    embed-resources: true
    slide-number: true
    scrollable: true
author: "Thi Mui Pham, Ph.D.<br>Jonathan Chipman, Ph.D.<br>PHS 7045 Teaching Team"
---

```{r}
#| echo: false
#| label: setup
#| include: false
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, results = "hold",
                      fig.width = 7, fig.height = 5)
```

## The two slogans of R

::: {.r-fit-text}

> To understand computations in R, two slogans help:
>
> - Everything that **exists** is an object.
> - Everything that **happens** is a function call.
>
> — John Chambers

These two sentences organize the whole day:

| | Chambers' slogan | What we do with it |
|---|---|---|
| **Part I** | everything that *exists* is an **object** | modes, structures, names, filtering, and what "vectorized" really means |
| **Part II** | everything that *happens* is a **function call** | promises, closures, functionals, non-standard evaluation |

:::

## The road map

::: {.r-fit-text}

**Part I: objects**

1. **Modes and structures**: when to use which one
2. **Naming and filtering**: get in the habit of using names, not positions
3. **Control flow**: how to write loops, and when you want one
4. **Functions**: how to write them, and how to handle the awkward cases
5. **Vectorization**: what is really vectorized, what only looks like it, and
   what actually makes R slow

**Part II: function calls**

6. **Everything is a function call**: operators, `if`, and `[` are functions
7. **Functions are objects**: store them, pass them, return them
8. **Lazy evaluation**: arguments are *promises*, not values
9. **Closures**: functions carry (and can change) their environment
10. **Functionals**: where `*apply` finally pays off
11. **Non-standard evaluation**: how `dt[, temp]` and `filter(df, x > 1)` work

::: {.fragment .fade-up}
Lazy evaluation (#8) is the idea everything else in Part II depends on. Force a
promise and you get a value; look at its expression instead and you get NSE.
:::

:::

## Preparation

::: {.r-fit-text}

Read/Watch: Selections from [R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/)

* Light read: Chapters 1–3
* Careful read: Chapters 4, 9–10, and 13
* Video: [Subsetting objects](https://www.youtube.com/watch?v=VfZUZGUgHqg)
* Video: [Vectorized Operations](https://www.youtube.com/watch?v=YH3qtw7mTyA)

Read: Selections from [The Art of R Programming](https://diytranscriptomics.com/Reading/files/The%20Art%20of%20R%20Programming.pdf#page=1)

* [Contents through Chapter 1](https://diytranscriptomics.com/Reading/files/The%20Art%20of%20R%20Programming.pdf#page=9) (light read, for familiarity)
* [Chapter 2: Vectors](https://diytranscriptomics.com/Reading/files/The%20Art%20of%20R%20Programming.pdf#page=51) (careful read)
* [Chapter 3, sections 5–end: Matrices and arrays](https://diytranscriptomics.com/Reading/files/The%20Art%20of%20R%20Programming.pdf#page=104)
* [Chapter 4, sections 1–3 and 5: Lists](https://diytranscriptomics.com/Reading/files/The%20Art%20of%20R%20Programming.pdf#page=104)
* [Chapter 7, sections 3–4](https://diytranscriptomics.com/Reading/files/The%20Art%20of%20R%20Programming.pdf#page=165)

Note: the big-picture principles in Chapter 2 (vectors) carry over to matrices,
lists, and data frames.

Aside: [a reference for good coding practices](https://www.r-bloggers.com/2018/09/r-code-best-practices/)

:::

# Part I: Everything that exists is an object

## Four ways to organize a program

::: {.r-fit-text}

Before looking at R's objects, it helps to know what kind of language R is. A
programming *paradigm* is a way of thinking about what a program **is**.
Paradigms differ in what they put at the center:

| Paradigm | Organized around | The program is… | Languages built around it |
|---|---|---|---|
| **Imperative / procedural** | steps that change stored values | a list of instructions: *how* to do it | C, Python, R |
| **Object-oriented** | objects that hold data together with the operations on it | objects calling each other's operations, updating what they hold | Java, C++, Python, R |
| **Functional** | functions, and combining them | expressions to work out, leaving stored values alone | Haskell, Lisp, R |
| **Logic / declarative** | facts, rules, and goals | *what* is true; the engine works out *how* | Prolog, SQL, R |

::: {.fragment .fade-up}
**A word on "stored values".** At any moment during a run, your variables hold
particular values. Programmers call that the program's *state*, and a variable
whose value can be changed after it is created is called *mutable*.

`x <- x + 1` changes the state: the name `x` now points at something different,
and whatever it pointed at before is gone. That is the imperative style. The
functional style avoids it: instead of changing `x`, you compute a new value and
give it its own name. Nothing that already exists is disturbed.
:::

::: {.fragment .fade-up}
**Notice that R appears in every row.** That is not a fudge: you will write
loops, you will use S3 classes, you will pass functions to `lapply()`, and you
will write `lm(y ~ x)`, which states a model rather than a fitting procedure.
These are lenses, not boxes. The useful question is never "which kind of
language is this", but which lens a given problem makes *natural*.
:::

::: {.fragment .fade-up}
**Why is Python not in the functional row?** Because the last column lists
languages *built around* a lens, not every language that permits it. Python has
first-class functions, `lambda`, `map()`, and closures, so you can certainly
write functionally in it. But it was not designed that way: values are mutable by
default, `reduce()` was demoted out of the builtins in Python 3, and Guido van
Rossum has explicitly refused tail-call optimisation, so deep recursion is not a
safe idiom. Idiomatic Python reaches for a comprehension or a loop.

R is in that row because its semantics were taken from Scheme: functions are
values, arguments are evaluated lazily, and every operation, `+` included, is a
function call. That is the difference between *supporting* a style and being
*built on* it.
:::

::: {.fragment .fade-up}
**One warning about the word "object".** R's slogan "everything that exists is
an object" means something weaker than the paradigm does: it says every value is
a thing you can name, store, and pass around. That alone is not
object-oriented programming.

R earns its place in that row separately, through S3, S4, and R6, which let you
attach a class to an object and have functions such as `summary()` behave
differently depending on it. We use S3 on the next slide but one.
:::

:::

## One task, four paradigms

::: {.r-fit-text}

Definitions only get you so far. Here is one task written four ways. (It is also
the task you will see again in the lab.)

**The task:** given trial outcomes, report the success rate of each arm.

```{r}
#| label: paradigm-data
#| output-location: fragment
set.seed(1)
trial <- data.frame(
  arm     = rep(c("control", "A", "B"), each = 10),
  success = rbinom(30, 1, 0.35)
)
head(trial, 3)
```

::: {.fragment .fade-up}
All four give the same numbers. What changes is what the code talks about.
:::

:::

## Imperative vs. functional

::: {.r-fit-text}

**Imperative.** You describe every step and change the stored values as you go:
`n`, `y`, and `rate` all hold something different after each pass. The code talks
about counters and positions.

```{r}
#| label: paradigm-imperative
#| output-location: fragment
arms <- unique(trial$arm)
rate <- numeric(length(arms))
names(rate) <- arms

for (a in arms) {
  n <- 0
  y <- 0
  for (i in seq_len(nrow(trial))) {     # walk every row, by hand
    if (trial$arm[i] == a) {
      n <- n + 1
      y <- y + trial$success[i]
    }
  }
  rate[a] <- y / n                      # change the stored result
}
rate
```

**Functional.** You combine small functions, and nothing that already exists is
changed: `split()` hands back new pieces, `vapply()` hands back a new vector. The
code talks about the transformation, split then average, rather than about
bookkeeping.

```{r}
#| label: paradigm-functional
#| output-location: fragment
vapply(split(trial$success, trial$arm), mean, numeric(1))
```

::: {.fragment .fade-up}
Same answer, fourteen lines versus one. The functional version never says *how*
to iterate; `split()` and `vapply()` take care of that. So there is no index to
get wrong and no counter to forget to reset.

(The arms come back in a different order, because `split()` sorts them. This is
a good reason to pull results out by name.)
:::

:::

## Object-oriented vs. declarative

::: {.r-fit-text}

**Object-oriented.** You bundle the data with the operations that belong to it,
and the object's class decides which code runs. The code talks about a trial
rather than about vectors.

```{r}
#| label: paradigm-oo
#| output-location: fragment
# S3: attach a class, then define a method for it
trial_obj <- structure(trial, class = c("trial", "data.frame"))

success_rate <- function(x, ...) UseMethod("success_rate")

success_rate.trial <- function(x, ...) {
  vapply(split(x$success, x$arm), mean, numeric(1))
}

success_rate(trial_obj)
```

Note that the method body is just the functional version. The object-oriented
part is not a different algorithm; it is a different way of organizing it and of
deciding which code to run.

**Declarative.** You say what you want and let something else work out how. No
iteration appears at all:

```sql
-- SQL: no loop, no counters, no iteration at all
SELECT arm, AVG(success) AS rate
FROM trial
GROUP BY arm;
```

::: {.fragment .fade-up}
R does this too. `lm(y ~ x, data = df)` states a model, not a fitting procedure,
and so does `aggregate(success ~ arm, trial, mean)`.
:::

:::

## What is R? Three lenses at once

::: {.r-fit-text}

R is object-oriented, functional, and a scripting language, all at once. Each
lens tells you something about the code you are about to write.

| Lens | What it means in R | Where it shows up today |
|---|---|---|
| **Object-oriented** | everything you make is an object: it knows what kind of data it holds, it can carry extra labels such as names or dimensions, and it can be labelled with what sort of thing it is | Part I: modes, structures, names |
| **Functional** | functions are values, and you combine them instead of writing the iteration by hand | Part I: vectorization. Part II: all of it |
| **Scripting** | a script runs top to bottom, interactively or in batch, and is reproducible | every simulation you run in this course |

::: {.fragment .fade-up}
**Class and generic.** That last label, saying what sort of thing an object is,
is called its **class**. The output of `lm()` has class `"lm"`; a data frame has
class `"data.frame"`.

A function that behaves differently depending on the class of its input is
called a **generic**. `summary()` is one: hand it a data frame and you get a
summary of each column, hand it the output of `lm()` and you get a regression
table. Same function name, different work, chosen by the class.
:::

::: {.fragment .fade-up}
R has more than one system for this. **S3** is the informal one you meet most
often, and the only one we use today. **S4** is the stricter, formal one, and
both ship with R. **Reference classes** (`setRefClass`) also come with R, and the
**R6** package is the popular alternative to them. We come back to these later in
the course.

R Manual, if you want the reference version:
[objects](https://cran.r-project.org/doc/manuals/r-release/R-intro.html#Objects),
[writing your own functions](https://cran.r-project.org/doc/manuals/r-release/R-intro.html#Writing-your-own-functions),
[scripting with R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html#Scripting-with-R).
Inspect any object with `str()` and `attributes()`.
:::

:::

## What makes R a scripting language

::: {.r-fit-text}

The difference is **when your code is turned into instructions the machine can
run**.

**Compiled (C, C++, Fortran).** You write source code, then a separate program,
the compiler, translates *the whole file* into machine code before anything
runs. You get an executable, and you run that.

```bash
gcc sim.c -o sim     # translate once, ahead of time
./sim                # then run the machine code
```

**Interpreted / scripting (R, Python, shell).** There is no build step. R reads
your code and carries it out as it goes, one expression at a time. The program
doing that reading is called the **interpreter**, and when we say "R" does
something at run time, that is what we mean. What you ship *is* the source file.

```bash
Rscript sim.R        # R reads and executes the file line by line
```

:::

## Consequences of scripting languages

::: {.r-fit-text}

Two consequences, both of which we meet later:

::: {.fragment .fade-up}
**1. R decides what your code means while it runs.** The compiler for C can
check the whole program up front and refuse to build it. R cannot: a typo on
line 200 is discovered when line 200 runs, twenty minutes into your simulation.
So run your code early and often, on a small number of replicates, before
scaling up.
:::

::: {.fragment .fade-up}
**2. Working things out at run time costs time.** For every operation, R has to
look up what the names mean and what type of data it is holding *right then*.
A compiled program settled all of that before it started. This is the whole
reason a loop written in R is slower than the same loop written in C, which we
come back to under vectorization.
:::

:::

# Data modes and structures

## Modes

::: {.r-fit-text}

An object's **mode** is the kind of data it holds: numbers, text, `TRUE`/`FALSE`,
and so on. Ask with `mode()` or `typeof()`.

```{r}
#| label: mode-demo
#| output-location: fragment
mode(1)
mode("a")
mode(TRUE)
```

Mode matters because it decides what you are allowed to do (you can add numbers,
not text) and how much memory each element takes. R has six primitive modes.

Two you will almost never create in practice:

* **Raw** (raw bytes, as they sit in memory): `x <- raw(2)` makes two bytes, both
  zero, printed as `00 00`. You would use this for reading binary files.
* **Complex** (numbers with an imaginary part): `x <- 0i` is the complex number
  0 + 0i. The `i` suffix is what makes it complex, as in `3 + 2i`.

One you create moderately often, usually indirectly:

* **Integer** (whole numbers): `x <- 1L`. The `L` suffix is what makes it an
  integer rather than a double. Without it, `x <- 1` gives you a numeric. R uses
  integers behind the scenes when you write `1:10`, and for counts and indices.

Three you create all the time:

* **Logical** (`TRUE`/`FALSE`): `x <- TRUE`
* **Numeric**, also called double (any real number): `x <- 0`
* **Character**, also called string: `x <- "hello world"`

::: {.fragment .fade-up}
Why `L` for integer? It is borrowed from C, where `L` means a "long" integer.
And why does `1` default to a double rather than an integer? Because most
numbers in statistics are not whole, so R makes the common case the default.
:::

:::

## Structures

::: {.r-fit-text}

| **Feature** | **Vectors** | **Matrices** | **Lists** | **Data frames** |
|---|---|---|---|---|
| **Memory usage** | Low | Low to moderate | Moderate | Moderate to high |
| **Dimensionality** | 1D | 2D (rows, columns) | 1D (elements may vary in dimension) | 2D (rows, columns) |
| **Homogeneity** | Homogeneous | Homogeneous | Heterogeneous | Heterogeneous by column |
| **Creation** | `c()`, `seq()`, `rep()`, `:` | `matrix()`, `cbind()`, `rbind()` | `list()` | `data.frame()` |
| **Combining** | `c()` | `cbind()`, `rbind()` | `c()`, `append()` | `cbind()`, `rbind()` |
| **Indexing** | `[ ]` | `[ , ]` | `[[ ]]`, `[ ]` | `$`, `[[ ]]`, `[ ]` |
| **Arithmetic** | Yes | Yes | No (operate on elements) | Not directly |
| **Advantages** | Memory-efficient, fast | Efficient for matrix math | Highly flexible | Optimized for tabular data |
| **Disadvantages** | Single data type | Single data type | Higher memory for complex elements | Attribute overhead |
| **Usage** | Simple collections | Mathematical computation | Mixed-type / complex structures | Tabular data |

:::

## Atomic modes

::: {.r-fit-text}

The six primitive modes are also called **atomic** modes: elements stored
together in a vector must share one mode.

So what happens when you mix them? R **coerces** everything to a single mode.

**Question:** what does `x` hold after this, and what is its mode?

```{r}
#| label: coercion-question
#| eval: false
x <- c("A", 0, 1L, TRUE)
x
str(x)
```

::: {.fragment .fade-up}
```{r}
#| label: coercion
#| echo: false
x <- c("A", 0, 1L, TRUE)
x
str(x)
```

Everything became character.
:::

::: {.fragment .fade-up}
The order, from least to most general, is:

`logical` → `integer` → `numeric` (double) → `character`

The most general mode wins, since it is the only one that can hold all the
values without losing information. `c(TRUE, 1L)` is integer, `c(1L, 2.5)` is
double, and anything mixed with a string becomes character.

R does this silently. It is a common source of bugs: one stray `"NA"` string
turns a whole numeric column into text.
:::

:::

## Missing values

::: {.r-fit-text}

R has several kinds of missing value, from most to least primitive:

* `NULL`: no value at all, has length 0
* `NaN`: not a number (for example `0/0`)
* `NA`: no information, has length 1
* `Inf` / `-Inf`: infinite

Companion tests: `is.null()`, `is.nan()`, `is.na()`, `is.infinite()`, and
`is.finite()` (which is `FALSE` for `NA`, `NaN`, and `Inf` alike).

```{r}
#| label: missing
#| output-location: fragment
x <- c(1, NA, NaN, Inf)
is.na(x)
is.finite(x)
```

Note that `is.na(NaN)` is `TRUE`: `NaN` is a special case of `NA`.

:::

## Vectors and matrices

::: {.r-fit-text}

Why to use vectors and matrices? They use less memory than lists and data
frames, and operations on them are faster. Get in the habit of using them
wherever your data are all one mode.

**Creating vectors.** To initialize an empty vector, use `vector()`,
`rep(NA, <length>)`, `numeric(<length>)`, or `character(<length>)`.

```{r}
#| label: create-vectors
#| output-location: fragment
vector("character", 2)
vector("numeric", 2)
rep(NA, 2)
numeric(2)
```

**Question:** why would you want to create an empty vector like this, rather
than just building the result up as you go?

::: {.fragment .fade-up}
Because you are reserving the memory in advance. If you know you need 1000
results, `numeric(1000)` asks R for one block of the right size once, and the
loop then fills in slots that already exist.

The alternative, starting from nothing and appending each result, makes R build
a new, slightly longer vector every time and copy everything into it. For 1000
results that is 1000 copies.

We measure how bad this gets when we come to vectorization.
:::

:::

---

::: {.r-fit-text}

Other common ways to build numeric vectors: `c()`, `:`, `seq()`, `rep()`.

```{r}
#| label: create-vectors-2
#| output-location: fragment
c(1, 2, 3, 4)          # combine
x <- 1:10              # sequential
seq(1, 10, by = 2)     # by increment
seq(1, 10, length.out = 3)   # equally spaced, fixed length
rep(1:2, each = 5)
rep(1:2, times = 5)

out <- rep(NA, 10)     # pre-allocate space for future results
```

**Creating matrices:** `matrix()`, `cbind()`, `rbind()`.

```{r}
#| label: create-matrices
#| output-location: fragment
x <- 1:10
rbind(x, x, x)
matrix(x, nrow = 2)
```

:::

## Naming and indexing

::: {.r-fit-text}

Elements are extracted by position, `[<position(s)>]`, or by name,
`[<name(s)>]`.

Rule of thumb: prefer names. Positions are transparent only to whoever wrote
them, and they shift the moment the data change.

```{r}
#| label: index-vector
#| output-location: fragment
x <- 1:10
x[3]
x[5:3]

names(x) <- 2011:2020
x[c("2015", "2017")]
```

```{r}
#| label: index-matrix
#| output-location: fragment
x <- matrix(1:10, nrow = 2)
rownames(x) <- 2010:2011
colnames(x) <- c("SLC", "Murray", "Bountiful", "Milcreek", "Sandy")
x
x[, c("SLC", "Milcreek")]

# Beware of dimension reduction
dim(x[, c("SLC", "Milcreek")])
dim(x["2010", c("SLC", "Milcreek")])
```

**Why is the last `dim()` `NULL`?**

::: {.fragment .fade-up}
Selecting a single row gives back a plain vector, and a vector has no `dim`
attribute. By default R drops dimensions of length 1. To keep them, use
`drop = FALSE`: `x["2010", c("SLC", "Milcreek"), drop = FALSE]` stays a 1 × 2
matrix. Watch for this inside functions and loops, where the change of shape is
easy to miss.
:::

:::

## Questions 1: Naming objects

::: {.r-fit-text}

1. What is the alphabet position of each letter of your name? Use a named
   vector (try the `LETTERS` object). For example, JONATHAN maps to
   10, 15, 14, 1, 20, 8, 1, 14.

2. Why is it important to extract elements through naming conventions?

:::

## Lists

::: {.r-fit-text}

Why to use lists? They are how a function returns several different kinds of
result at once. `lm()` returns a list, of class `lm`, which is why `print()` and
`summary()` treat it specially.

```{r}
#| label: create-lists
#| output-location: fragment
x <- list(1:2, "A", NULL, c(TRUE, FALSE), list(1:10, "B"))
str(x)

x <- list("2010" = 1, "2011" = "A", "2013" = TRUE)
names(x)
```

Extract with `[ ]` (returns a list), `[[ ]]` (returns the element), or `$`.
See [subsetting lists](https://bookdown.org/rdpeng/rprogdatascience/subsetting-r-objects.html#subsetting-lists).

```{r}
#| label: index-lists
#| output-location: fragment
x <- list("2010" = 1:2, "2011" = "A", "2013" = c(TRUE, FALSE))
x[c("2010", "2011")]   # a list of two elements
x[["2010"]]            # the element itself
x$"2010"
```

:::

## Data frames

::: {.r-fit-text}

Why to use data frames?

* They are the workhorse of data analysis.
* But they are slower to compute with. Selecting `d[i, j]` from a data frame is
  several times slower than `m[i, j]` from a matrix, because `[.data.frame` is R
  code that has to work out what you meant, while `[` on a matrix is a primitive.
  For heavy computation, use a matrix unless you need mixed types.
* Underneath, a data frame is a list of equal-length columns, so list indexing
  works on it.

A helpful data frame comes from `expand.grid()`, which builds every combination
of the values you give it. How can this be helpful in simulations?

```{r}
#| label: expand-grid
#| output-location: fragment
expand.grid("Param 1" = 1:2, "Param 2" = letters[1:3])
```

::: {.fragment .fade-up}
You get every simulation scenario in one object, one row per scenario. Then you
iterate over its rows instead of writing nested loops, and your results line up
row for row with the parameter grid.
:::

:::

## `data.table` and `tibble`

::: {.r-fit-text}

A `data.table` **is a data frame**: it carries the `data.frame` class as well as
its own, so anything written for data frames still works on it. The same holds
for the tidyverse's `tibble`. They are data frames with extra behaviour, not
replacements.

```r
library(data.table)
dt <- data.table(temp = c(1, 2, 3), site = c("a", "a", "b"))

class(dt)          # "data.table" "data.frame"
is.data.frame(dt)  # TRUE
nrow(dt)           # 3 — ordinary data frame functions are fine
```

What you get for using one, in practice:

- **Grouped summaries are much faster.** On 5 million rows,
  `aggregate(temp ~ site, df, mean)` takes about 0.76 seconds;
  `dt[, mean(temp), by = site]` takes about 0.02, roughly 36 times faster. The
  gap widens with the number of groups.
- **Reading files is dramatically faster.** A 25 MB csv takes `read.csv()` about
  0.85 seconds and `fread()` about 0.013, and `fread()` guesses column types
  more sensibly.
- **Columns can be added and changed without copying the table.** More on this
  on the next slide.
- **Joins are keyed and fast**, rather than going through `merge()` on unsorted
  columns.

:::

## Why `data.table` is fast

::: {.r-fit-text}

**Filter, compute, and group happen in one pass.** For a data frame, `df[i, j]`
selects rows and columns and nothing else. For a data.table, the same brackets
take a filter, an expression to compute, and a grouping:

```r
dt[temp > 1, mean(temp), by = site]
```

Because that is one call, `data.table` can plan the whole operation: it never
builds the filtered copy that `df[df$temp > 1, ]` would build before summarising
it.

::: {.fragment .fade-up}
**Columns are modified in place.** Adding a column to a data frame duplicates
the table object; `data.table` writes into the existing one:

```r
df$new  <- df$temp * 2      # data frame: the object is duplicated
dt[, new := temp * 2]       # data.table: no duplication at all
```

`tracemem(df)` prints a line each time R duplicates the object, and you get two
for the data frame and none for the data.table.

Be careful how much you read into that. The duplication is *shallow*: R copies
the list of columns, not the values inside them, so adding one column to a
5-million-row data frame still takes only a few milliseconds. It matters when
you add or modify columns repeatedly, and in the loops where that adds up.
:::

::: {.fragment .fade-up}
**Bare column names.** `temp` is not a variable in your workspace, so `[` cannot
simply evaluate it. It captures the expression instead. We come back to how that
works at the end of Part II.
:::

:::

## Questions 2: Structures

::: {.r-fit-text}

1. What structure would you want for a computationally heavy task?

2. What structure would you want for reading in data from a colleague?

3. What kind of object does the following return?

```{r}
#| label: q1-lm
y <- 1:10
x <- 1:10
f <- lm(y ~ x)
```

What do you think `mode(f)` and `names(f)` will give you?

```{r}
#| label: q1-lm-inspect
#| eval: false
mode(f)
names(f)
```

:::

# Naming conventions and filtering

## Naming your objects

::: {.r-fit-text}

Earlier we named the *elements* of an object so we could pull them out by name.
This is the other kind of naming: what you call the objects themselves.

Borrowing from a [blog on R code best practices](https://www.r-bloggers.com/2018/09/r-code-best-practices/),
there are five naming conventions in the wild:

* `alllowercase`: e.g. `adjustcolor`
* `period.separated`: e.g. `plot.new`
* `underscore_separated`: e.g. `numeric_version`
* `lowerCamelCase`: e.g. `addTaskCallback`
* `UpperCamelCase`: e.g. `SignatureMethod`

Strive for names that are concise and meaningful:

* Avoid a name that would overwrite something that already exists. For example,
  `c <- 0` overwrites the `c()` function.
* Clarity is better than brevity unless it is well documented. What does the `s`
  in `s <- 0` stand for: sample, simulation, success?

Naming conventions are personal preference. Pick one and stay with it. This
course uses `underscore_separated`.

:::

## Filtering vectors

::: {.r-fit-text}

Elements can be filtered by position, by name, or by Boolean logic
(`==`, `!=`, `>`, `>=`, `<`, `<=`).

Example: 10 patients randomized to control (0) or treatment (1).

```{r}
#| label: filter-1
#| output-location: fragment
tx <- rep(0:1, times = 5)
names(tx) <- 1:10

tx[tx == 1]
tx[c(1, 3, 7)]
tx["3"]
```

Note that `tx[tx == 1]` does two things at once: `==` returns a vector of
`TRUE`/`FALSE`, and `[` selects with it. Both are function calls, which we come
back to in Part II.

Boolean logic also works with `which()` (which positions meet the criterion) and
`any()` / `all()` (do any or all of them meet it):

```{r}
#| label: filter-2
#| output-location: fragment
which(tx == 1)
any(tx == 1)
all(tx == 1)
```

:::

## Questions 3: Filtering objects

::: {.r-fit-text}

Suppose `x` is the number of vacations taken in each year 2010–2019.

```{r}
#| label: q3-setup
#| output-location: fragment
set.seed(1)
x <- sample.int(n = 7, size = 10, replace = TRUE)
names(x) <- 2010:2019
x
```

1. How many vacations were taken in the first and ninth years?

2. How many total vacations were taken in odd years? Use element names.

3. How many total vacations were taken in even years? Use `seq()`.

4. Why would you want to use element names whenever possible?

:::

# Control statements

## If-then statements

::: {.r-fit-text}

> "R is a block-structured language ... delineated by braces, though braces are
> optional if the block consists of just a single statement. Statements are
> separated by newline characters or, optionally, by semicolons."
> — Matloff, [page 139](https://diytranscriptomics.com/Reading/files/The%20Art%20of%20R%20Programming.pdf#page=165)

```{r}
#| label: if-else
#| output-location: fragment
x <- "yes"
if (x == "yes") {
  print("I'll take on the project")
} else {
  print("Sorry, I can't take on the project")
}
```

An `if`/`else` statement needs a single `TRUE`/`FALSE`. `ifelse()` applies the
same conditional to every element of a vector:

```{r}
#| label: ifelse
#| output-location: fragment
x <- 1:10
ifelse(test = x > 5, yes = 1, no = 0)
```

:::

## Loops

::: {.r-fit-text}

Loops repeat an operation over the values in a vector:

* `for`: iterate through each value of a vector
* `while`: continue while a condition is `TRUE`
* `repeat`: continue until a `break` or `return` statement

```{r}
#| label: loops-for
#| output-location: fragment
x <- c(3, 7, 11)
for (n in x) print(n)             # iterate over VALUES

for (i in seq_along(x)) print(x[i])   # iterate over POSITIONS
```

Use `seq_along(x)` rather than `1:length(x)`. If `x` is empty, `1:length(x)` is
`c(1, 0)` and the loop runs twice over nothing, while `seq_along(x)` does not run
at all.

**What will the following print?**

```{r}
#| label: loops-quiz
#| eval: false
x <- seq(1, 10, by = 3)
for (i in x) print(i)
```

::: {.fragment .fade-up}
`1`, `4`, `7`, `10`. The loop variable takes the *values* in `x`, not the
positions.
:::

:::

---

::: {.r-fit-text}

Use `while` and `repeat` when you do not know in advance how many iterations you
need:

```{r}
#| label: loops-while
#| output-location: fragment
i <- 1
while (i <= 10) i <- i + 4
i

i <- 1
repeat {
  i <- i + 4
  if (i > 10) break
}
i
```

`break` exits the loop; `next` skips to the next iteration.

::: {.fragment .fade-up}
A loop is the imperative lens again: you describe every step and update a
running result. In Part II we see the functional alternative, where you hand the
body of the loop to another function and let it do the iterating. Neither is
automatically better, and knowing when to use which is the point.
:::

:::

## Questions 4: Creating a loop

::: {.r-fit-text}

From Jan 1–9, 2023, it snowed 6 days atop Snowbasin. Each day is 1 (snow) or
0 (no snow).

```{r}
#| label: q4-setup
snow <- c(1, 1, 1, 1, 0, 1, 1, 0, 0)
names(snow) <- 1:9
```

On which day did it *first* snow three consecutive days (that day plus the two
before)? Solve with a loop and conditional logic.

:::

## Functions

::: {.r-fit-text}

This is where the functional lens becomes concrete. Functions reduce redundancy
and make code clearer. Matloff
([page xxii](https://diytranscriptomics.com/Reading/files/The%20Art%20of%20R%20Programming.pdf#page=23))
on using functions instead of copy-paste:

* Clearer, more compact code
* Potentially faster execution
* Less debugging, because the code is simpler
* Easier transition to parallel programming

```{r}
#| label: fn-examples
#| output-location: fragment
fun1 <- function(v1, v2) v1 + v2       # implicit return of last expression
fun1(1, 2)

fun2 <- function(v1, v2) {
  out <- c(sum = v1 + v2, ave = (v1 + v2) / 2)
  return(out)                          # a named vector
}
fun2(1, 2)

fun3 <- function(v1, v2) {
  list(sum = v1 + v2, ave = (v1 + v2) / 2, text = "fun3")  # a list, for mixed types
}
str(fun3(1, 2))
```

:::

---

::: {.r-fit-text}

A function with a default argument, for the "unfair coin experiment":

```{r}
#| label: unfair-coin
#| output-location: fragment
# unfairCoin
#   n: number of tosses
#   p: probability of heads (default = 0.7)
unfairCoin <- function(n, p = 0.7) {
  sample(c("H", "T"), n, replace = TRUE, prob = c(p, 1 - p))
}

set.seed(1)
tosses <- unfairCoin(20)
prop.table(table(tosses))
```

Defaults let you write `unfairCoin(20)` for the common case and
`unfairCoin(20, p = 0.5)` when you need something else.

::: {.fragment .fade-up}
Two things here are stranger than they look, and Part II explains both. The
default `p = 0.7` is not evaluated until it is used. And `unfairCoin` is itself
just an object with a name attached, no different from `tosses`.
:::

:::

## Questions 5: Creating a function

::: {.r-fit-text}

Generalize your Question 4 code into a function so that, for any string of days,
you can find the first day it snowed a given number of consecutive days. Return
`NA` if no day meets the criterion.

:::

# Vectorizing and replicate

## What "vectorized" really means

::: {.r-fit-text}

R can perform the same operation on every element of a vector at once. A
vectorized function takes the whole vector and does the looping in C, in one
pass, without going back to the R interpreter for each element.

Both versions loop. The difference is *who* does the looping, and the next slide
explains why that matters so much.

::: {.fragment .fade-up}
So "vectorized" is a claim about **how a function is built**. Plenty of
functions accept a whole vector and hand back a whole vector while still looping
in R, one element at a time. They look vectorized from the outside and give you
none of the speed. When we say a function is *genuinely* vectorized, we mean the
loop is really in C.
:::

```{r}
#| label: vec-basics
#| output-location: fragment
x1 <- 1:10
x2 <- 101:110
x1 + x2        # element-wise: element i of one, with element i of the other
```

:::

## Recycling: the rule that makes it work

::: {.r-fit-text}

A vectorized operation pairs up elements, which leaves an obvious question: what
if the two vectors are not the same length?

R's answer is **recycling**. It repeats the shorter operand until the lengths
line up. That is what makes a comparison against a single number mean anything
at all:

```{r}
#| label: vec-recycle-1
#| output-location: fragment
x1 <- 1:10
x1 > 5         # the single 5 is repeated to length 10, then compared
```

Without recycling you would have to write `x1 > rep(5, 10)` every time.

::: {.fragment .fade-up}
The caution is that R does this **silently** whenever the longer length is a
multiple of the shorter:

```{r}
#| label: vec-recycle-2
#| output-location: fragment
x <- 1:6
y <- 1:3
x + y          # y is used twice, no warning at all
```

You only get told when the lengths do not divide evenly:

```{r}
#| label: vec-recycle-3
#| output-location: fragment
#| warning: true
1:6 + 1:4      # warning: longer object length is not a multiple...
```

So the dangerous case is the tidy one. A filter that accidentally returns 3
values instead of 6 will be recycled without complaint.
:::

:::

## What counts as vectorized

::: {.r-fit-text}

Genuinely vectorized: the arithmetic and comparison operators, `sqrt()`,
`log()`, `exp()`, `abs()`, `round()`, `ifelse()`, `pmin()` and `pmax()`,
`cumsum()`, `sum()`, `mean()`, `which()`, `is.na()`, and the random number and
distribution functions (`rbinom()`, `rbeta()`, and so on).

:::

## How can you tell?

::: {.r-fit-text}

**Look at the function.** Print it. If the body is `.Primitive`, `.Internal`, or
`.Call`, the work is being handed straight to compiled code:

```{r}
#| label: vec-tell
#| output-location: fragment
sqrt                # .Primitive
body(paste)         # .Internal
body(stats::rnorm)  # .Call
```

**Or just time it against a loop.** If a function is really doing the work in C,
it should be far faster than the same thing written out element by element. If
the times are similar, the loop is still in R somewhere.

::: {.fragment .fade-up}
A good example of something that only *looks* vectorized. `Vectorize()` accepts
vectors and returns a vector, but underneath it is `mapply()`, which calls your
function once per element:

```{r}
#| label: vec-fake
#| output-location: fragment
slow_max <- Vectorize(function(a, b) if (a > b) a else b)

x <- runif(2e5)
system.time(pmax(x, 0.5))[["elapsed"]]        # real vectorization, in C
system.time(slow_max(x, 0.5))[["elapsed"]]    # a loop in disguise
```

Same interface, same answer, roughly a hundredfold difference.
:::

:::

## What `.Primitive`, `.Internal` and `.Call` mean

::: {.r-fit-text}

All three say the same thing: **the work leaves the R interpreter and runs as
compiled C code**. They differ in who is allowed to use them and how much R
wrapping sits in front.

| Marker | What it is | Who can use it |
|---|---|---|
| **`.Primitive`** | there is no R function at all. The name is bound straight to C. The most basic operations work this way: `+`, `[`, `sum`, `c`, `sqrt` | base R only |
| **`.Internal`** | a normal R function that checks arguments and fills in defaults, then hands the real work to a C routine in R's own internal table | base R only |
| **`.Call`** | runs a C function that someone wrote, compiled, and loaded into your session. The C function receives your R objects, loops over them at C speed, and returns a new R object. This is the one you will write yourself later, via `Rcpp` | anyone |

```r
sqrt                 # function (x)  .Primitive("sqrt")
body(paste)          # .Internal(paste(list(...), sep, collapse, recycle0))
body(stats::rnorm)   # .Call(C_rnorm, n, mean, sd)
```

::: {.fragment .fade-up}
Why so many? Historical, mostly. `.Primitive` is fastest, since there is no R
function call and no argument matching first, but it is reserved for the handful
of operations R cannot do without. `.Internal` gives base R a readable R layer in
front of C.

For today the practical reading is simply: **if you see any of them, the
per-element loop is not happening in R.**
:::

:::

## Which ones are not?

::: {.r-fit-text}

None of the functions on the previous list. Those all do their per-element work
in C. The ones that *look* vectorized but are not fall into two groups.

**A `for` loop written in R.** You can read the loop in the function's own
source: `apply()` and `Reduce()`.

**The loop is in C, but it calls back into R every element.** `lapply()`,
`sapply()`, `vapply()`, `mapply()`, and `Map()` hand the iteration to
`.Internal(...)`, so the counter is not the slow part. The cost is that the
compiled loop must return to the interpreter to evaluate **your** R function for
each element, and that is exactly the overhead vectorization exists to avoid.
`Vectorize()` wraps `mapply()`, `replicate()` wraps `sapply()`, and `Filter()`
wraps `lapply()`, so all three inherit it.

::: {.fragment .fade-up}
The distinction rarely matters in practice: both groups pay R-level cost per
element, and both are far behind a genuinely vectorized call.
:::

```{r}
#| label: vec-not
#| output-location: fragment
m <- matrix(runif(2e6), ncol = 10)

system.time(rowMeans(m))[["elapsed"]]        # C loop over the whole matrix
system.time(apply(m, 1, mean))[["elapsed"]]  # R loop, calling mean() 200,000 times
```

::: {.fragment .fade-up}
**One caution.** Vectorized does not automatically mean fast. `ifelse()` is
genuinely vectorized, but it builds several intermediate vectors and is still a
few times slower than `pmax()` for the same job. C-level looping removes the
interpreter overhead; it does not excuse doing unnecessary work.
:::

:::

## Why the loop being in C matters

::: {.r-fit-text}

R is an **interpreted** language: nothing is translated into machine code ahead
of time. Each time R evaluates `out[i] <- sqrt(x[i])`, it works out afresh what
that line means. Roughly, per iteration, R must:

1. look up `sqrt`, `x`, `out`, and `i` by name. R searches the current
   **environment**, which is just the table of names and values a piece of code
   can see, then the one enclosing it, and so on out to the packages you have
   loaded;
2. check what `x` is (a double vector? an integer? does it have a class with its
   own `[` method?) and pick the right code;
3. build the argument as a promise, and allocate a small object to hold `x[i]`;
4. check for `NA` and for the result being out of range;
5. assign into `out`, checking whether `out` is shared with anything else and
   needs copying first.

That bookkeeping costs far more than the square root itself.

::: {.fragment .fade-up}
`sqrt(x)` does step 1 and 2 **once**, then hands the whole vector to a compiled C
loop. That loop was translated to machine code when R was built. It already
knows the values are doubles sitting next to each other in memory, so the
per-element cost is one machine instruction plus a check for `NA`, with no
lookups and no allocation.

So a vectorized call is not doing less work in total. It is paying the
interpreter's overhead once instead of `n` times.
:::

:::

## The `*apply` family is not vectorization

::: {.r-fit-text}

This is the part of the usual advice that is out of date. `sapply()`,
`lapply()`, `vapply()`, `mapply()`, and `purrr::map()` are loops written in C
that call your R function once per element. The R work still happens one element
at a time.

```{r}
#| label: vec-apply-vs-loop
#| output-location: fragment
x <- runif(1e5)

# a loop, pre-allocated
out1 <- numeric(length(x))
for (i in seq_along(x)) out1[i] <- sqrt(x[i])

out2 <- vapply(x, sqrt, numeric(1))    # a loop wearing a functional's clothes
out3 <- sqrt(x)                        # ACTUALLY vectorized

all.equal(out1, out3); all.equal(out2, out3)
```

::: {.fragment .fade-up}
On 5 million elements the loop takes about 1.4 seconds and `vapply()` about 0.5,
so the functional is a few times quicker: it skips the R-level `out1[i] <- ...`
on every pass. But `sqrt(x)` takes about 0.005 seconds, which is roughly a
hundred times faster again.

That is the point. `*apply` and a `for` loop are in the same league, because both
evaluate R code once per element. Genuine vectorization is in a different one.

So use `*apply` and `purrr` because they are clearer, not because they close that
gap. 
:::

:::

## What actually makes R code slow

::: {.r-fit-text}

Since R 3.4, R quietly translates your functions into a compact internal form
the first time they run, which takes some of the interpreter overhead out of
loops. A `for` loop is not slow in itself. Here is what is actually slow:

```{r}
#| label: vec-grow
#| output-location: fragment
#| cache: false
n <- 20000

grow <- function(n) {                # BAD: reallocates and copies each step
  out <- c()
  for (i in 1:n) out <- c(out, i^2)
  out
}

prealloc <- function(n) {            # GOOD: allocate once, fill in place
  out <- numeric(n)
  for (i in 1:n) out[i] <- i^2
  out
}

vectorized <- function(n) (1:n)^2    # BEST, when the operation is vectorized

system.time(grow(n))[["elapsed"]]
system.time(prealloc(n))[["elapsed"]]
system.time(vectorized(n))[["elapsed"]]
```

::: {.fragment .fade-up}
`grow()` gets slower and slower as it goes: every `c()` allocates a new vector
and copies everything into it. That, rather than the `for` keyword, is the
classic R performance bug, and it is why we set up the result vector in advance.

The copying is not arbitrary. R vectors are copy-on-modify, which comes straight
out of the functional lens: do not change something other code might be holding.
Part II shows the one kind of object that opts out, the environment.
:::

:::

## A better set of rules of thumb

::: {.r-fit-text}

The old advice was "always vectorize, never loop". Here is a more useful set of
rules:

1. **Write it clearly first.** Correct and readable beats fast and wrong.
2. **Never grow an object inside a loop.** Set it up first with `numeric(n)`,
   `vector("list", n)`, or `matrix()`, or collect the pieces and `rbind()` them
   once at the end.
3. **Use a genuinely vectorized function when one exists.** `sqrt(x)`,
   `rbinom(n, ...)`, and `colSums(m)` are free speed.
4. **Use `*apply` and `purrr` for clarity**, and because they are easy to run in
   parallel later. In code you plan to keep, use `vapply()` rather than
   `sapply()`: you say what shape the output should be, and you get an error if
   it is not.
5. **Keep the loop when each step depends on the last.** Adaptive randomization
   and MCMC cannot be vectorized away, and forcing them hurts readability for
   nothing.
6. **Measure before you optimize.** `bench::mark()` and `profvis::profvis()`. We
   spend a whole week on this.

```r
library(bench)
bench::mark(loop = prealloc(n), vec = vectorized(n), relative = TRUE)
```

::: {.fragment .fade-up}
And when a loop that really is sequential is still too slow, the answer is not a
cleverer `apply`. It is Rcpp or parallel computing, both later in this course.
:::

:::

## `replicate()` for simulation

::: {.r-fit-text}

`replicate()` evaluates an expression `n` times and collects the results. It
calls `sapply()` underneath, so it is a loop rather than vectorization, but it is
the natural way to write a Monte Carlo experiment.

```{r}
#| label: replicate
#| output-location: fragment
set.seed(1)
tosses <- replicate(1000, sum(sample(0:1, 10, replace = TRUE)))
mean(tosses)
```

Watch the `simplify` argument. `replicate()` returns a vector or matrix when it
can and a list when it cannot, which is the same sometimes-surprising behavior as
`sapply()`.

Note also that it takes an expression, not a value, and re-evaluates it each
time. That is lazy evaluation, which we get to in Part II.

:::

# Part II: Everything that happens is a function call

## Where Part I left off

::: {.r-fit-text}

Part I was about the first slogan: everything that exists is an object, with a
mode, attributes, and a class. Along the way we left three questions open.

- Why is `sapply()` no faster than a loop?
- Why is growing a vector with `c()` so slow, when changing an environment
  costs nothing?
- How does `dt[, temp]` find a column that is not a variable in your workspace?

::: {.fragment .fade-up}
All three are questions about what happens when a function is called, which is
the second slogan, and the rest of today.

1. **Everything is a function call**: operators, `if`, and `[` are functions.
2. **Functions are objects**: stored, passed, returned, sometimes unnamed.
3. **Lazy evaluation**: arguments are promises, not values.
4. **Scoping and closures**: functions carry their environment, and environments
   can be changed.
5. **Functionals**: where `*apply` finally pays off.
6. **Non-standard evaluation**: capturing code instead of values.

Number 3 is the one the rest depends on. Everything before it explains how
arguments and scopes behave, and NSE is what you get when you capture a promise
instead of forcing it.
:::

:::

# 1. Everything is a function call

## Operators, `if`, and `[` are all functions

::: {.r-fit-text}

We opened with Chambers' second slogan: everything that happens is a function
call. Take it literally. Operators, `if`, subsetting, and even assignment are
function calls in disguise.

```{r}
#| label: fn-call-1
#| output-location: fragment
`+`(2, 3)                  # `+` is an ordinary function
`if`(2 > 1, "yes", "no")   # if returns a value
`[`(c(10, 20, 30), 2)      # subsetting is a function too
```

::: {.fragment .fade-up}
So `x[i]` in your loop, `x + y` in your vectorized code, and `tx[tx == 1]` when
you filtered the treatment assignments were all function calls.
:::

:::

---

::: {.r-fit-text}

Because operators are just functions, **you can define your own**:

```{r}
#| label: fn-call-2
#| output-location: fragment
`%+%` <- function(a, b) paste0(a, b)
"data" %+% ".table"
```

::: {.fragment .fade-up}
Why this matters: if even operators and `if` are functions, then functions are
nothing special. They are values, and you can store them, pass them, and return
them. Everything else in Part II builds on that.
:::

:::

# 2. Functions are objects

## First-class functions

::: {.r-fit-text}

A function is a value like any other. You can put it in a list, pass it as an
argument, or return it from another function. (When a language lets you do this
with functions, they are called **first-class**.)

```{r}
#| label: first-class-1
#| output-location: fragment
funs <- list(
  root   = sqrt,
  square = function(x) x^2,
  neg    = function(x) -x
)
funs$square(5)
```

```{r}
#| label: first-class-2
#| output-location: fragment
# A function that takes a function and returns a function
apply_twice <- function(f) function(x) f(f(x))
add1 <- function(x) x + 1
apply_twice(add1)(10)
```

:::

---

::: {.r-fit-text}

**Anonymous functions** have no name. Since R 4.1 there is a shorthand, `\(x)`:

```{r}
#| label: anon
#| output-location: fragment
(\(x) x^2)(4)          # same as (function(x) x^2)(4)
```

::: {.fragment .fade-up}
We will lean on anonymous functions constantly once we reach *functionals*.
:::

:::

# 3. Lazy evaluation

## Arguments are promises, not values

::: {.r-fit-text}

In most languages every argument is computed before the call runs. R is
different. An argument is a *promise*: the expression you wrote, plus the
environment to evaluate it in. R only evaluates it when the function first uses
it.

```{r}
#| label: lazy-1
#| output-location: fragment
f <- function(x, y) x     # note: y is never used
f(1, stop("boom"))        # no error: y is never forced!
```

::: {.fragment .fade-up}
**Why would this go wrong elsewhere?** Because most languages are *eager*: they
evaluate each argument expression at the call site, before the function body
starts. Whether the parameter is used inside makes no difference, because the
function has not been entered yet when the evaluating happens.

What that costs you depends on the language. In Python, the equivalent raises
before `f` ever runs:

```python
def f(x, y): return x        # y is never used
f(1, boom())                 # ValueError - boom() ran at the call site
```

C has no exceptions, but the argument is still evaluated, so any side effect or
wasted computation happens anyway:

```c
int f(int x, int y) { return x; }   /* y unused */
f(1, boom());                       /* boom() still runs */
```

In R nothing ever asks for `y`, so the promise is never evaluated and
`stop("boom")` never runs.
:::

:::

## Question 6: What does this print? {.smaller}

::: {.r-fit-text}

```r
g <- function(x, y) {
  x * 2
}
g(5, stop("this never runs"))
```

:::

## Consequence: defaults can see other arguments

::: {.r-fit-text}

A default is a promise evaluated inside the function, so it can refer to the
other arguments. Most languages cannot do this.

```{r}
#| label: lazy-default
#| output-location: fragment
summarise_vec <- function(x, n = length(x)) {
  sprintf("mean of %d values: %.2f", n, mean(x))
}
summarise_vec(c(2, 4, 6))    # n defaults to length(x)
```

:::

## The classic gotcha: closures in a loop

::: {.r-fit-text}

Each function below captures the promise for `i`, and that promise is not
evaluated until later, after the loop has already finished.

```{r}
#| label: lazy-gotcha
#| output-location: fragment
make_fns <- function() {
  fns <- vector("list", 3)
  for (i in 1:3) fns[[i]] <- function() i
  fns
}
fns <- make_fns()
c(fns[[1]](), fns[[2]](), fns[[3]]())
```

::: {.fragment .fade-up}
`force(i)` evaluates the promise right away. `lapply()` gives each iteration its
own `i`, so it avoids the problem altogether:

```{r}
#| label: lazy-fix
#| output-location: fragment
fns2 <- lapply(1:3, function(i) function() i)
c(fns2[[1]](), fns2[[2]](), fns2[[3]]())  # 1 2 3
```
:::

:::

# 4. Scoping & closures

## Lexical scoping: functions carry their environment

::: {.r-fit-text}

A function remembers where it was *defined*, not where it is called. It keeps
hold of the names that were visible at that moment, and a function together with
those kept names is called a **closure**.

```{r}
#| label: closure-factory
#| output-location: fragment
# A "function factory": a function that manufactures functions
power <- function(exp) {
  function(x) x^exp
}
square <- power(2)
cube   <- power(3)
c(square(4), cube(2))
```

::: {.fragment .fade-up}
`square` and `cube` have the same body. Each one kept a different `exp`.
:::

:::

## Environments can be changed in place

::: {.r-fit-text}

Vectors and lists are
copy-on-modify: change one and R gives you back a modified copy, leaving the
original alone. That is the functional style built into the language, and it is
why `grow()` got slower and slower.

Environments are the exception. Hand one to a function and the function changes
the original, because what gets passed is a pointer to the environment rather
than a copy of its contents.

```{r}
#| label: env-counter
#| output-location: fragment
# `<<-` assigns in the enclosing environment, so the value survives between calls
new_counter <- function() {
  n <- 0
  function() {
    n <<- n + 1
    n
  }
}
count <- new_counter()
c(count(), count(), count())    # 1 2 3, the closure remembers n
```

```{r}
#| label: env-ref
#| output-location: fragment
e <- new.env(); e$value <- 1
modify <- function(env) env$value <- 99   # no copy is made
modify(e)
e$value                                    # 99, modified in place
```

:::

## Aside: replacement functions

::: {.r-fit-text}

Here is one more place the second slogan turns up. Assigning into a function
call is itself a function call.

```{r}
#| label: replacement
#| output-location: fragment
x <- c(1, 2, 3)
names(x) <- c("a", "b", "c")   # secretly calls `names<-`(x, value)
x
```

`names(x) <- v` is really ``x <- `names<-`(x, v)``. The same holds for
`dim(x) <-`, `levels(x) <-`, and `x[i] <-`: there is a function behind each one,
and you can write your own.

Note that this does *not* mean a copy is made. `names<-` is a primitive, so if
nothing else points at `x`, R modifies it in place and rebinds the name. A copy
happens only when the vector is shared, which is copy-on-modify doing its job.

:::

# 5. Functionals

## Functions that take functions

::: {.r-fit-text}

A *functional* takes a function as an input and applies it over your data,
replacing an explicit loop with a single call.

```{r}
#| label: func-lapply
#| output-location: fragment
lapply(1:3, function(x) x^2)      # apply over a list, get a list back
```

```{r}
#| label: func-vapply
#| output-location: fragment
vapply(1:3, function(x) x^2, numeric(1))   # you state the shape of each result
```

::: {.fragment .fade-up}
The advantage of `vapply()` compared to `sapply()`: it gives an
error instead of quietly returning the wrong type. 
:::

:::

## More functionals: Map, Reduce, Filter

::: {.r-fit-text}

```{r}
#| label: func-map
#| output-location: fragment
Map(function(x, y) x + y, 1:3, 10:12)   # iterate over several inputs at once
```

```{r}
#| label: func-reduce
#| output-location: fragment
Reduce(`+`, 1:5)                        # combine the elements two at a time -> 15
Reduce(`+`, 1:5, accumulate = TRUE)     # keep the running total
```

```{r}
#| label: func-filter
#| output-location: fragment
Filter(function(x) x %% 2 == 0, 1:10)   # keep the elements where this is TRUE
```

The tidyverse offers a type-stable family via **purrr**:

```r
library(purrr)
map_dbl(1:3, \(x) x^2)        # 1 4 9  (like vapply, always a double vector)
map2_dbl(1:3, 10:12, `+`)     # 11 13 15
```

:::

## Why we care: functionals are the bridge to parallelism

::: {.r-fit-text}

Once your iteration is written as a functional, running it in parallel is a
one-line change, because the function is just an object you can hand to a
different engine.

```r
# today
lapply(files, process)

# in a few weeks: same shape, now parallel / on the cluster
parallel::mclapply(files, process, mc.cores = 4)
future.apply::future_lapply(files, process)
```

::: {.fragment .fade-up}
This only works because functions are objects. It is also the answer to the
question from Part I: `*apply` is not faster on its own, but this is where the
speed eventually comes from.
:::

:::

# 6. Non-standard evaluation

## Capturing code instead of values

::: {.r-fit-text}

Lazy evaluation, taken one step further. Instead of evaluating a promise to get
its value, a function can look at the expression itself with `substitute()`.

```{r}
#| label: nse-1
#| output-location: fragment
g <- function(x) deparse(substitute(x))
g(a + b * 2)     # "a + b * 2"; a and b need not exist
```

```{r}
#| label: nse-standard-vs-not
#| output-location: fragment
h_standard <- function(x) x              # sees the VALUE
h_nse      <- function(x) substitute(x)  # sees the CODE
h_nse(1 + 1)                             # `1 + 1`, unevaluated
```

This is *non-standard evaluation* (NSE): the function reads the code you typed
rather than the value it produces.

:::

## You have been using this all along

::: {.r-fit-text}

Every time you type a bare, unquoted name, NSE is at work:

```r
subset(df, temp > 20)            # `temp` is not a variable in your workspace...
dplyr::filter(df, temp > 20)     # ...it is a bare name, captured then evaluated
dt[, .(temp, wind)]              # data.table does exactly this
lm(y ~ x, data = df)             # y ~ x is a captured expression (a formula)
library(dplyr)                   # bare name, not the string "dplyr"
```

None of these would work under standard evaluation. R would look for `temp` in
the global environment and report "object not found".

::: {.fragment .fade-up}
Compare this with Part I, where you pulled things out with a name you had
attached yourself, as in `x["2010"]`. NSE is the same idea built into the
language: the column name is the thing you type.
:::

:::

## Build your own `subset()` in two lines

::: {.r-fit-text}

The whole trick is to capture the expression and then evaluate it somewhere of
your choosing, here inside the data frame.

```{r}
#| label: nse-subset
#| output-location: fragment
my_subset <- function(data, condition) {
  cond <- substitute(condition)          # 1. capture the code, unevaluated
  rows <- eval(cond, envir = data)       # 2. evaluate it INSIDE `data`
  data[rows, , drop = FALSE]
}
head(my_subset(mtcars, mpg > 30))        # `mpg` resolves to a column, not a global
```

:::

## The catch: NSE is awkward to program around

::: {.r-fit-text}

NSE is convenient when you type at the console and awkward when you write
functions around it. Watch this quietly wrong result:

```{r}
#| label: nse-catch
#| output-location: fragment
my_var <- "mpg"
nrow(my_subset(mtcars, my_var > 30))   # 32, ALL rows!
nrow(mtcars)
```

`my_var` was captured as code, not as the string it holds, so the comparison used
the name itself. No error, no warning, and the wrong answer.

::: {.fragment .fade-up}
Getting out of this, that is, telling the function to evaluate one argument
normally, is what the tidyverse's `{{ }}` and `.data[[ ]]`, and data.table's
`env =`, are for.
:::

:::

# Wrap-up

## What to take away

::: {.r-fit-text}

**Everything that exists is an object**

1. **Structures.** Vectors and matrices for computation, lists and data frames
   when the pieces differ.
2. **Names.** Pull things out by name, not by position. This is the habit that
   pays off most.
3. **Control flow.** Loops are fine. Growing an object inside one is not.
4. **Functions.** Turn the fixed numbers into arguments, handle the awkward
   cases, and return as soon as you have the answer.
5. **Vectorization.** Real vectorization happens in C. `*apply` is a loop, and
   you use it because it reads better.

**Everything that happens is a function call**

6. **Operators and `if` are functions.** Functions are not a special case.
7. **Functions are objects.** Store them, pass them, return them.
8. **Lazy evaluation.** Arguments are promises, evaluated when first used.
9. **Closures.** A function carries its environment and can change it. This is
   the exception to the copying that made point 3 slow.
10. **Functionals.** Apply a function over your data. This is where the clarity
    of point 5 turns into speed, once you run it in parallel.
11. **NSE.** Take the promise's expression instead of its value.

::: {.fragment .fade-up}
In one sentence: objects are what R has, function calls are what R does, and
lazy evaluation sits between them.
:::

:::

## Where this goes next

::: {.r-fit-text}

- **Debugging and profiling:** measure before you optimize, with `bench` and
  `profvis`.
- **Parallel computing:** your `lapply()` becomes `mclapply()` or
  `future_lapply()`.
- **R packages:** closures, `...`, and functionals are what a clean interface is built from.

:::

## Session info {.smaller}

```{r}
#| label: sessioninfo
sessionInfo()
```
