---
title: "Building R packages (part I)"
subtitle: "PHS 7045: Advanced Programming"
author: "George G. Vega Yon, Ph.D."
format:
  revealjs:
    embed-resources: true
    slide-number: true
    scrollable: true
---

# Intro

```{r}
#| echo: false
#| label: setup
#| include: false
knitr::opts_chunk$set(eval = TRUE, fig.width = 7, fig.height = 5)
slides_eval <- TRUE
```

## Today 

::: {style="font-size: 20pt"}

  - Have a look at some R-package-dev practices based primarily on my own
    experience (so the product is sold AS-IS)

  - We won’t discuss either Rcpp or coding R itself. (for that I
    recommend [“Advance R”](http://adv-r.had.co.nz/Rcpp.html), and
    there’s also Patrick Muchmore’s [“A template for R packages using
    Rcpp and/or
    RcppArmadillo”](https://github.com/patrickmuchmore/RPackageTemplate))

  - We won’t discuss GitHub in depth. I’ll assume that the user already
    knows the basics of git(hub): Creating a repo online and start
    using it locally. (A good reference is [“Pro
    Git”](https://git-scm.com/book/en/v2))

  - We won’t discuss software engineering/design; we’ll 
    look at some tools and dev cycles. (For that, you have [“Code
    Complete”](http://cc2e.com/))

:::


## What we’ll be using today

::: {style="font-size: 20pt"}

1.  **devtools**: “Package development tools for R” ([read
    more](https://cran.r-project.org/package=devtools)).

2.  **roxygen2**: “A ‘Doxygen’-like in-source documentation system for
    Rd, collation, and ‘NAMESPACE’ files” ([read
    more](https://cran.r-project.org/package=roxygen2)).

3.  **covr**: (An R package for) “Test Coverage for Packages” ([read
    more](https://cran.r-project.org/package=covr)).

4.  **RStudio**: “RStudio is an integrated development environment (IDE)
    for R” ([read more](https://cran.r-project.org/package=covr)).

5.  **valgrind**: (if you use C/C++ code) “a memory error detector”
    ([read more](http://valgrind.org/)).

:::


## Why ‘spend’ time writing an R package?

::: {style="font-size: 20pt"}

To name a few:

1.  Easy way to share code: Type `install.packages` and voilà\!

2.  Already standardized: So you don’t need to think about how to
    structure it.

3.  CRAN checks everything for you: Force yourself to code things right.

4.  **Reproducible Research**
    
      - Accessible Reproducible Research ([Mesirov, **Science**
        2010](http://science.sciencemag.org/content/327/5964/415))
    
      - Again, and Again, and Again, ... ([Jasny et al., **Science**
        2011](http://science.sciencemag.org/content/334/6060/1225))
    
      - Challenges in Irreproducible Research ([**nature**
        topic](http://www.nature.com/news/reproducibility-1.17552))

:::

## What’s an R package, anyway?

According to Hadley Wickham’s [“R packages”](http://r-pkgs.had.co.nz/)

> Packages are **the fundamental units of reproducible R code**. They
> include **reusable R functions**, the **documentation** that describes
> how to use them, and **sample data**.

<div id="htmlwidget-85ed755bf6d9f6419a39" class="d3wordcloud html-widget" style="width:600px;height:576px;">

## What’s an R package, anyway? (cont.) 

::: {style="font-size: 20pt"}

Folders

  - **R** Where the R scripts live, e.g., `addnums.r`, `funny-pkg.r`.

  - **tests** Where the tests live, e.g., `testthat/test-addnums.r`
    (using testthat).

  - **vignettes** Where vignettes live, e.g.,
    `vignettes/mode_details_on_addnums.rmd`.

  - **data** Where the data lives, e.g., `fakedata.rda`,
    `aneattable.csv`.

  - **man** Where the manuals live, e.g., `addnums.Rd`, `funnypkg.Rd`.

And files

  - **DESCRIPTION** The metadata of the package: name, description,
    version, author(s), dependencies, etc.

  - **NAMESPACE** Declares the name of the objects (functions mainly)
    that will be part of the package (and will be visible to the user)

More info [here](http://r-pkgs.had.co.nz/package.html) and
[here](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Package-structure)

:::


## Dev cycle

::: {style="font-size: 20pt"}

Writing R packages is an iterative process

1.  **Set up the structure**: Create folders and files `R/`, `data/`,
    `src/`, `tests/`, `man/`, `vignettes/`, `DESCRIPTION`,
    `NAMESPACE`(?)

2.  **Code\!** For `f` in `F` do
    
    1.  Write `f`
    
    2.  Document `f`: what it does, what the inputs are, what it
        returns, and examples.
    
    3.  Add some tests: is it doing what it is supposed to do?
    
    4.  `R CMD Check`: Will CRAN and the rest of the functions ‘like’
        it?
    
    5.  Commit your changes\!: Update `ChangeLog`+`news.md`, and commit
        (so travis and friends run)\!
    
    6.  next `f`

3.  Submit your package to CRAN/Bioconductor (read
    [more](http://r-pkgs.had.co.nz/release.html#release-submission))

:::

# Handson

## Step 1: Set up the structure

::: {style="font-size: 20pt"}

You have several methods: `package.skeleton()`, Rstudio’s “New Project”,
`devtools::new()`, etc. I personally do it from scratch (including the
code):

1.  Create the package dir, say `funnypkg`

2.  Create the `R` dir

3.  Create the `DESCRIPTION` file ([read
    more](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#The-DESCRIPTION-file)):
    
    ``` yml
    Package: funnypkg
    Type: Package
    Title: What the Package Does (Title Case)
    Version: 0.1.0
    Author: Who wrote it
    Maintainer: The package maintainer <yourself@somewhere.net>
    Description: More about what it does (maybe more than one line)
        Use four spaces when indenting paragraphs within the Description.
    License: MIT + file
    Encoding: UTF-8
    LazyData: true
    ```

4.  From RStudio, create a new project pointing to the package folder.

:::

## Step 1: Set up the structure (cont. 1)

::: {style="font-size: 20pt"}

And, using the `devtools` package, we can add some extras

```r
# Creating a README for the project
devtools::use_readme_rmd()

# Infrastructure for testing
devtools::use_testthat()

# Infrastructure for Code Coverage
devtools::use_coverage(type = "codecov") # This creates the .travis.yml
devtools::use_appveyor()

# A LICENSE file (required by CRAN)
# The line "license: MIT + file" in the DESCRIPTION file
devtools::use_mit_license(copyright_holder = "George G. Vega Yon")
devtools::use_news_md()
```

:::

## Step 1: Set up the structure (cont. 2)

::: {style="font-size: 20pt"}

What did happen?

1.  `use_readme_rmd()`: Creates a readme file that will be in the project's main
    folder. Think about it as the home page of your project.

2.  `use_testthat()`: The basic infrastructure for testing packages will be created.

3.  `use_coverage(type = "codecov")`: Creates the [.travis.yml
    file](https://en.wikipedia.org/wiki/Comparison_of_continuous_integration_software)
    for Unix CI, and sets it up for code coverage using
    [codecov.io](https://codecov.io/).[^travis]

4.  `devtools::use_appveyor()`: Creates the
    [appveyor.yml](https://www.appveyor.com/) for windows CI.

5.  `devtools::use_mit_license(copyright_holder = "George G. Vega
    Yon")`: Creates the—- `LICENSE` file and puts it under ‘George G.
    Vega Yon’.

6.  `devtools::use_news_md()`: Creates the `news.md` file which us used
    for tracking changes and communicating them to the users (e.g.
    [netdiffuseR](https://github.com/USCCANA/netdiffuseR/blob/master/NEWS.md))

Now that we have the structure, we can start coding!

[^travis]: Today, the most common way to set up continuous integration is via GitHub actions.
We will learn more about that next week.

:::

## Step 2.1 and 2.2: Write `f` and Document it

::: {.columns style="font-size: 20pt"}

::: {.column width="40%"}

- Using `roxygen2` is very straight forward. For our fist pice of
    code, we create the `R/addnums.r` file (right).

- Type `devtools::document()`, or press Ctrl + Shift + D (RStudio
    will: create the manual and the NAMESPACE). Make sure you activate
    this option in RStudio (not the default).

- Notice that the output is defined using S3-type objects ([read
    more](http://adv-r.had.co.nz/S3.html))

:::

::: {.column width="55%"}

  
    
```r
#' The title of -addnums-
#'
#' Here is a brief description
#' 
#' @param a Numeric scalar. A brief description.
#' @param b Numeric scalar. A brief description.
#' 
#' @details Computes the sum of \code{x} and \code{y}.
#' @return A list of class \code{funnypkg_addnums}:
#' \item{a}{Numeric scalar.}
#' \item{b}{Numeric scalar.}
#' \item{ab}{Numeric scalar. the sum of \code{a} and \code{b}}
#' @examples
#' foo(1, 2)
#' 
#' @export
addnums <- function(a, b) {
    ans <- a + b
    structure(list(a = a, b = b, ab = ans)
    , class = "funnypkg_addnums")
}
```

:::

:::

## Step 2.1 and 2.2: Write `f` and Document it (cont. 1)

::: {.columns style="font-size: 20pt"}

::: {.column width="55%"}

```r
#' @rdname addnums
#' @export
#' @param x An object of class \code{funnypkg_addnums}.
#' @param y Ignored.
#' @param ... Further arguments passed to
#' \code{\link[graphics:plot.window]{plot.window}}.
plot.funnypkg_addnums <- function(x, y = NULL, ...) {
    graphics::plot.new()
    graphics::plot.window(xlim = range(unlist(c(0,x))), ylim = c(-.5,1))
    graphics::axis(1)
    with(x, graphics::segments(0, 1, ab, col = "blue", lwd=3))
    with(x, graphics::segments(0, 0, a, col = "green", lwd=3))
    with(x, graphics::segments(a, .5, a + b, col = "red", lwd=3))
    graphics::legend("bottom", col = c("blue", "green", "red"), 
            legend = c("a+b", "a", "b"), bty = "n", 
            ncol = 3, lty = 1, lwd=3)
}
```

:::

::: {.column width="40%"}

  - Here, we are adding a function that is documented in `addnums`. The
    plot method (left).

  - You’ll need to update the `man/*Rd` every time that you add new
    roxygen content. Press Ctrl + Shift + D, and RStudio will do it
    for you.

:::
:::

::: {style="font-size: 20pt"}

 - Notice that functions from foreign packages are called using the
    `::` operator (read more
    [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/ns-dblcolon.html)
    and [here](http://r-pkgs.had.co.nz/namespace.html)). Not a
    requirement, but makes the code easier to maintain (and less
    error-prone... because of R’s scoping rules).

:::

## Step 2.1 and 2.2: Write `f` and Document it (cont. 2)

::: {.columns style="font-size: 20pt"}

Since we added foreign functions, we need to add them to the NAMESPACE
and to the DESCRIPTION files:

::: {.column width="50%"}

`R/funny-pkg.r`

```r
#' @importFrom graphics plot.new plot.window axis legend
NULL

#' funnypkg
#'
#' A (not so) funny collection of functions
#'
#' @description We add stuff up... You can access to the project
#' website at \url{https://github.com/USCbiostats/software-dev/tree/master/rpkgs/funnypkg}
#'
#' @docType package
#' @name funnypkg
#'
#' @author George G. Vega Yon
NULL
```

:::

::: {.column width="50%"}

`DESCRIPTION`

```yml
Package: funnypkg
Type: Package
Title: What the Package Does (Title Case)
Version: 0.1.0
Author: Who wrote it
Maintainer: The package maintainer <yourself@somewhere.net>
Description: More about what it does (maybe more than one line)
    Use four spaces when indenting paragraphs within the Description.
License: What license is it under?
Encoding: UTF-8
LazyData: true
Suggests:
    testthat,
    covr
RoxygenNote: 5.0.1
Imports: graphics
```
  
:::

The `Suggests` and `RoxygenNote` fields were added automagically by
    `devtools`.

:::

## Step 2.3: Add some tests

::: {style="font-size: 20pt"}

Test are run every time that you run R CMD check

  - In the `tests/testthat/` dir, add/edit a source file with tests,
    e.g. `test-addnums.r`
    
    ```r
    context("Basic set of tests")
    test_that("addnums(a, b) = a+b", {
      # Preparing the test
      a <- 1
      b <- -2
    
      # Calling the function
      ans0 <- a+b
      ans1 <- addnums(a, b)
    
      # Are these equal?
      expect_equal(ans0, ans1$ab)
    })
    
    test_that("Plot returns -funnypkg_foo-", {
      expect_s3_class(plot(addnums(1,2)), "funnypkg_foo")
    })
    ```
  - You can run the tests using Ctrl + Shift + t

  - Furthermore, tests are used to evaluate code coverage.

:::

## Step 2.4: R CMD check

::: {style="font-size: 20pt"}

Running `R CMD check` is easy with RStudio

  - If you don’t have C/C++ code Just press Ctrl + Shift + E

  - If you have C/C++ code, use `R CMD Check` with valgrind (check for
    segfaults)
    
    ```shell
    $ R CMD build funnypkg/
    $ R CMD check --as-cran --use-valgrind funnypkg*.tar.gz
    ```
    
    You can ask RStudio to use valgrind too.

For an extensive list of the checks, see the section [1.3.1 Checking
packages](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Checking-packages)
in the “Writing R extensions” manual.


How does the output of `R CMD check` looks like?

```shell
==> devtools::check()

Updating funnypkg documentation
Loading funnypkg
Setting env vars ---------------------------------------------------------------
CFLAGS  : -Wall -pedantic
CXXFLAGS: -Wall -pedantic
Building funnypkg --------------------------------------------------------------
'/usr/lib/R/bin/R' --no-site-file --no-environ --no-save --no-restore --quiet  \
  CMD build  \
  '/home/vegayon/Dropbox/usc/core_project/CodingStandards/rpkgs/funnypkg'  \
  --no-resave-data --no-manual 

* checking for file ‘/home/vegayon/Dropbox/usc/core_project/CodingStandards/rpkgs/funnypkg/DESCRIPTION’ ... OK
* preparing ‘funnypkg’:
* checking DESCRIPTION meta-information ... OK
* checking for LF line-endings in source and make files
* checking for empty or unneeded directories
* building ‘funnypkg_0.1.999.tar.gz’

Setting env vars ---------------------------------------------------------------
_R_CHECK_CRAN_INCOMING_USE_ASPELL_: TRUE
_R_CHECK_CRAN_INCOMING_           : FALSE
_R_CHECK_FORCE_SUGGESTS_          : FALSE
Checking funnypkg --------------------------------------------------------------
'/usr/lib/R/bin/R' --no-site-file --no-environ --no-save --no-restore --quiet  \
  CMD check '/tmp/RtmpBYGSto/funnypkg_0.1.999.tar.gz' --as-cran --timings  \
  --no-manual 

* using log directory ‘/home/vegayon/Dropbox/usc/core_project/CodingStandards/rpkgs/funnypkg.Rcheck’
* using R version 3.3.2 (2016-10-31)
* using platform: x86_64-pc-linux-gnu (64-bit)
* using session charset: UTF-8
* using options ‘--no-manual --as-cran’
* checking for file ‘funnypkg/DESCRIPTION’ ... OK
* checking extension type ... Package
* this is package ‘funnypkg’ version ‘0.1.999’
* package encoding: UTF-8
* checking package namespace information ... OK
* checking package dependencies ... OK
* checking if this is a source package ... OK
* checking if there is a namespace ... OK
* checking for executable files ... OK
* checking for hidden files and directories ... OK
* checking for portable file names ... OK
* checking for sufficient/correct file permissions ... OK
* checking whether package ‘funnypkg’ can be installed ... OK
* checking installed package size ... OK
* checking package directory ... OK
* checking DESCRIPTION meta-information ... OK
* checking top-level files ... OK
* checking for left-over files ... OK
* checking index information ... OK
* checking package subdirectories ... OK
* checking R files for non-ASCII characters ... OK
* checking R files for syntax errors ... OK
* checking whether the package can be loaded ... OK
* checking whether the package can be loaded with stated dependencies ... OK
* checking whether the package can be unloaded cleanly ... OK
* checking whether the namespace can be loaded with stated dependencies ... OK
* checking whether the namespace can be unloaded cleanly ... OK
* checking loading without being on the library search path ... OK
* checking dependencies in R code ... OK
* checking S3 generic/method consistency ... OK
* checking replacement functions ... OK
* checking foreign function calls ... OK
* checking R code for possible problems ... OK
* checking Rd files ... OK
* checking Rd metadata ... OK
* checking Rd line widths ... OK
* checking Rd cross-references ... OK
* checking for missing documentation entries ... OK
* checking for code/documentation mismatches ... OK
* checking Rd \usage sections ... OK
* checking Rd contents ... OK
* checking for unstated dependencies in examples ... OK
* checking examples ... OK
* checking for unstated dependencies in ‘tests’ ... OK
* checking tests ...
  Running ‘testthat.R’ OK
* DONE
Status: OK




R CMD check results
0 errors | 0 warnings | 0 notes

R CMD check succeeded
```

:::

## Step 2.5: Commit your changes

::: {style="font-size: 20pt"}

Once you have run R CMD checks and no errors or unexpected issues are
detected, you should commit your changes and push them to Github.

1.  Edit the
    [`NEWS.md`](https://github.com/hadley/dplyr/blob/master/NEWS.md)
    file and the
    [`ChangeLog`](https://github.com/RcppCore/RcppArmadillo/blob/master/ChangeLog)
    (if you have one).

2.  Add the new files to the tree, commit your changes, and push them to
    Github
    
    ```shell
    $ git add R/newfile.r man/newfile.Rd
    $ git commit -a -m "Adding new function"
    $ git push
    ```

Once uploaded, I recommend checking Travis and AppVeyor to see if
everything went well.

:::

## Step 3: Submit your package to CRAN/Bioconductor

::: {style="font-size: 20pt"}


  - You can release your package using `devtools::release()` function.

  - The devtools package was built to make life easier. Such is it’s
    creator’s confidence in the `release` function that it comes with a
    warranty\!
    
    > If a devtools bug causes one of the CRAN maintainers to **treat
    > you impolitely**, I will personally send you a handwritten apology
    > note. Please forward me the email and your address, and I’ll get a
    > card in the mail.

:::

## Dev Cycle (once again\!)

::: {style="font-size: 20pt"}

The most important step

2.  **Code\!** For `f` in `F` do
    
    1.  Write `f`
    
    2.  Document `f`: what it does, what the inputs are, what it
        returns, examples.
    
    3.  Add some tests: is it doing what is supposed to do?
    
    4.  `R CMD Check`: Will CRAN, and the rest of the functions, ‘like’
        it?
    
    5.  Commit your changes\!: Update `ChangeLog`+`news.md`, and commit
        (so travis and friends run)\!
    
    6.  next `f`


:::

# Fin

```{r}
devtools::session_info()
```


## More resources

- [“Writing R
extensions”](https://cran.r-project.org/doc/manuals/r-release/R-exts.html)
(R Core Team)

- [“R packages”](http://r-pkgs.had.co.nz/) (Hadley Wickham)

- [“Developing Packages with
RStudio”](https://support.rstudio.com/hc/en-us/articles/200486488-Developing-Packages-with-RStudio)
(Josh Paulson)

- [“A template for R packages using Rcpp and/or
RcppArmadillo”](https://github.com/patrickmuchmore/RPackageTemplate)
(Patrick Muchmore)

- [“R package primer: a minimal
tutorial”](http://kbroman.org/pkg_primer/) (Karl Broman)

- [“Writing an R package from
scratch”](https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/)
(Hilary Parker)

- [“Instructions for Creating Your Own R
Package”](http://web.mit.edu/insong/www/pdf/rpackage_instructions.pdf)
(Kim *et al.*)



