---
title: "Lab 03 - R Essentials and Functions (Solutions)"
subtitle: "PHS 7045: Advanced Programming — Fall 2026"
format:
  html:
    embed-resources: true
---

```{r}
#| label: setup
#| message: false
set.seed(7045)
```

# Warm-up: loops, functions, and `beepr`

**A1. Play each sound in turn.**

```{r}
#| eval: false
for (k in 1:11) {
  beepr::beep(k)
  Sys.sleep(2)
}
```

**A2. Pause a random duration.** `runif(1, 1, 3)` gives a pause between one and
three seconds.

```{r}
#| eval: false
for (k in 1:11) {
  beepr::beep(k)
  Sys.sleep(runif(1, min = 1, max = 3))
}
```

**B.** A beep is an *out-of-band* signal: it tells you a long-running job
finished without you watching the console. That is only useful because R is a
scripting language, i.e., the code runs top to bottom in a live session, so side effects like sound, printing, and plotting are part of the normal workflow (compared to a compiled binary).

**C.** `beepr::beep()` says exactly which package the function came from. It
avoids masking (two attached packages exporting the same name), it documents the
dependency at the call site, and it keeps a script runnable even if someone
removes the `library()` call. The cost is verbosity and a very small lookup
overhead, so the usual compromise is `library()` for packages used constantly
and `::` for one-off calls.

**D. Cumulative sum by hand.**

```{r}
cumsum_beep <- function(x, sound = 1) {
  if (length(x) == 0L) {
    warning("`x` is empty; returning a 0-row result.", call. = FALSE)
    return(cbind(x = numeric(0), cumsum = numeric(0)))
  }
  if (!is.numeric(x)) {
    stop("`x` must be a numeric vector.", call. = FALSE)
  }
  out <- numeric(length(x))
  total <- 0
  for (i in seq_along(x)) {
    total <- total + x[i]
    out[i] <- total
  }
  if (requireNamespace("beepr", quietly = TRUE)) beepr::beep(sound)
  return(cbind(x = x, cumsum = out))
}

# Test if it behaves as intended for the edge case of an empty vector
x <- NULL
res <- cumsum_beep(x)

# Test for an actual vector
x <- c(3, 1, 4, 1, 5, 9, 2, 6)
res <- cumsum_beep(x)
res
# Check against cumsum()
all.equal(res[, "cumsum"], cumsum(x), check.attributes = FALSE)
```

Preallocating `out` matters: growing a vector with `out <- c(out, total)`
reallocates on every iteration, which turns a linear loop into a quadratic one.

Validating `x` first also matters. Without the checks, `cumsum_beep(NULL)`
skips the loop entirely (`seq_along(NULL)` is empty) but still falls through to
the beep -- a silent no-op that sounds like success. Note the two different
responses: a non-numeric `x` is a programming error, so `stop()`; an empty `x`
is a plausible edge case (a filter that matched nothing), so `warning()` plus an
early return of a 0-row result keeps the caller's pipeline running while still
making noise in the log. Check inputs before any side effect.

Order matters here: `is.numeric(NULL)` is `FALSE`, so testing `is.numeric()`
first would send `cumsum_beep(NULL)` down the error branch instead of the
warning branch. `length(NULL)` is `0`, so the emptiness check catches both `NULL`
and `numeric(0)` when it comes first.

# Part 1: The urn problem

## 1. A first version, then a function

The direct translation of the problem: draw one marble from urn 1, move it into
urn 2, draw from urn 2, record whether that second marble is blue.

```{r}
urn_prob_loop <- function(nreps = 100000, b1 = 10, y1 = 8, b2 = 6, y2 = 6) {
  blue <- logical(nreps)
  for (i in seq_len(nreps)) {
    # Draw from urn 1: TRUE if blue.
    first_blue <- runif(1) < b1 / (b1 + y1)

    # Transfer it into urn 2 and draw again.
    b2_now <- b2 + first_blue
    blue[i] <- runif(1) < b2_now / (b2 + y2 + 1)
  }
  mean(blue)
}

urn_prob_loop(nreps = 10000)
```

## 2. Check against the analytic answer

### Where the formula comes from

The formula in the lab is just the **law of total probability** applied to the
one thing that is uncertain about urn 2: the color of the marble transferred
into it.

Let $B_1$ be the event that the first draw (from urn 1) is blue, and $B_2$ the
event that the second draw (from urn 2) is blue. The two cases $B_1$ and
$B_1^c$ partition the sample space, so

$$
\Pr(B_2) = \Pr(B_2 \mid B_1)\Pr(B_1) + \Pr(B_2 \mid B_1^c)\Pr(B_1^c).
$$

The three ingredients:

- **The first draw.** Urn 1 holds $b_1$ blue and $y_1$ yellow marbles, all
  equally likely, so $\Pr(B_1) = \frac{b_1}{b_1 + y_1}$ and
  $\Pr(B_1^c) = \frac{y_1}{b_1 + y_1}$.

- **If a blue marble was transferred,** urn 2 now holds $b_2 + 1$ blue out of
  $b_2 + y_2 + 1$ total, so $\Pr(B_2 \mid B_1) = \frac{b_2 + 1}{b_2 + y_2 + 1}$.

- **If a yellow marble was transferred,** the blue count is unchanged but the
  total still went up by one:
  $\Pr(B_2 \mid B_1^c) = \frac{b_2}{b_2 + y_2 + 1}$.

Substituting gives exactly the expression in the lab,

$$
\Pr(B_2) = \frac{b_1}{b_1+y_1}\cdot\frac{b_2+1}{b_2+y_2+1}
         + \frac{y_1}{b_1+y_1}\cdot\frac{b_2}{b_2+y_2+1}.
$$

With the default urns, $\frac{10}{18}\cdot\frac{7}{13} + \frac{8}{18}\cdot\frac{6}{13} = \frac{118}{234} \approx 0.5043$.

Two things are worth noticing. First, the denominator $b_2 + y_2 + 1$ is the
same in both terms — urn 2 gains a marble either way — so only the *numerator*
records which color arrived. Second, the structure of the derivation is the
structure of the simulation: `first_blue` is the conditioning event, and the
`prob` argument of the second `rbinom()` call is the conditional probability
$\Pr(B_2 \mid \cdot)$. The simulation is doing the same bookkeeping, one
replicate at a time, and averaging instead of weighting.

### What "Monte Carlo error" means

The simulation answers a *deterministic* question — $\Pr(B_2)$ is a fixed number
— with a *random* procedure, so the answer is random too. Run
`urn_prob_loop()` twice and you get two different values. Monte Carlo error is
the size of that wobble: the discrepancy between $\hat p$ and $p$ that comes
from having used finitely many replicates, not from a bug.

It is worth separating this from the other way an estimate can be wrong:

- **Monte Carlo error** shrinks as `nreps` grows. It is noise.
- **Bias** does not. If the code transfers the marble incorrectly, more
  replicates only pin down the wrong number more precisely.

That distinction is the whole reason to quantify the error: it is the only way
to tell "my simulation disagrees with the truth in the third decimal because it
is noisy" from "my simulation disagrees because it is broken."

**Why the standard error has that form.** Each replicate is an independent trial
that comes up "second marble is blue" with probability $p$. Writing
$X_i \in \{0,1\}$ for replicate $i$, the estimate is the sample mean
$\hat p = \frac{1}{n}\sum_{i=1}^n X_i$ of $n = \texttt{nreps}$ i.i.d. Bernoulli
draws. A Bernoulli variable has $\mathrm{Var}(X_i) = p(1-p)$, and variances of
independent variables add, so

$$
\mathrm{Var}(\hat p) = \frac{1}{n^2}\sum_{i=1}^n \mathrm{Var}(X_i)
                     = \frac{p(1-p)}{n},
\qquad
\mathrm{SE}(\hat p) = \sqrt{\frac{p(1-p)}{n}}.
$$

Since $p$ is what we are trying to estimate, we plug in $\hat p$ for it. Two
consequences are worth internalizing:

- **The $\sqrt{n}$ law.** The error falls like $1/\sqrt{n}$, so one extra decimal
  place of accuracy costs **100×** the replicates. This is what makes
  brute-force simulation expensive and vectorized code worth writing.
- **The problem sets the difficulty.** $p(1-p)$ is largest at $p = 0.5$ — which
  is almost exactly our case — and small for rare events. A probability near
  one half is the hardest kind to pin down.

By the central limit theorem $\hat p$ is approximately normal for large $n$, so
$\hat p \pm 1.96\,\widehat{\mathrm{SE}}$ covers $p$ about 95% of the time. That
interval is the yardstick: "within Monte Carlo error" means the interval covers
the truth, not that the two numbers look close to the eye.

### The check

```{r}
urn_prob_exact <- function(b1 = 10, y1 = 8, b2 = 6, y2 = 6) {
  b1 / (b1 + y1) * (b2 + 1) / (b2 + y2 + 1) +
    y1 / (b1 + y1) *  b2      / (b2 + y2 + 1)
}

truth <- urn_prob_exact()
truth
```

Applying that yardstick:

```{r}
nreps <- 100000
phat  <- urn_prob_loop(nreps = nreps)
se    <- sqrt(phat * (1 - phat) / nreps)
lower <- phat - 1.96 * se
upper <- phat + 1.96 * se

c(estimate = phat, se = se, lower = lower, upper = upper, truth = truth)
```

**How to read that output.** Each element answers a different question:

- `estimate` is $\hat p$, our simulated answer.
- `se` is the Monte Carlo standard error: it says how much `estimate` would move
  if we re-ran the simulation with a different seed. Here it is about
  $0.0016$, i.e. only the first two decimals of `estimate` are trustworthy.
  Reporting `estimate` to six digits would be false precision.
- `lower`/`upper` are the 95% interval $\hat p \pm 1.96\,\widehat{\mathrm{SE}}$.
- `truth` is the analytic value.

**The verdict is a single yes/no question: does the interval `[lower, upper]`
cover `truth`?** Ask R rather than squinting at the numbers:

```{r}
covers <- (lower <= truth) && (truth <= upper)

c(distance = truth - phat, se = se, n_se_away = (truth - phat) / se)
covers
```

`truth - phat` is the actual error, about $1.3\times 10^{-4}$, and `se` says an
error of that size is entirely ordinary — it is well under one standard error,
so `covers` is `TRUE` and the function passes — the simulation and the algebra
agree to within Monte Carlo error. Note what we did *not* do: we never asked
whether `estimate` and `truth` "look close." Closeness only means something
relative to `se`.

That `n_se_away` column is the check as one number, and it is the version worth
automating:

```{r}
z <- (phat - truth) / se
abs(z) < 2
```

$z$ is how many standard errors the estimate sits from the truth, so under a
correct implementation it is approximately a standard normal draw. Roughly
$|z| < 2$ is unremarkable, $|z| > 3$ is a red flag. Written this way the check
is a one-line regression test: it does not care what `nreps` is, because the
yardstick rescales with the precision.

**What to do with the answer:**

- **The interval covers the truth ($|z|$ small).** The implementation is consistent
  with the analytic answer. If the interval is still too wide for the use you
  have in mind, that is a *precision* problem, not a correctness one — increase
  `nreps` (and remember the $1/\sqrt{n}$ law: ten times narrower costs 100×
  the replicates).
- **The interval misses the truth ($|z|$ large).** Do not reach for more
  replicates: bias does not shrink with `nreps`, so a bug will only be pinned
  down more precisely. Go looking for the bug instead. In this problem the usual
  culprits are forgetting to add the transferred marble to urn 2's denominator,
  or drawing the second marble from urn 1's composition.

Note the direction of that sentence. $p$ is a fixed, unknown constant — here we
even know it exactly — and it is the *interval* that is random, since both
`estimate` and `se` are recomputed from fresh draws on every run. So the
question is always whether this particular interval covered $p$, never whether
$p$ happened to fall inside fixed bounds. "95% confidence" is a property of the
procedure across hypothetical repetitions: about 95% of the intervals it
produces cover $p$. It says nothing about the one interval in front of you,
which either covers $p$ or does not.

That is also the caveat: about 5% of intervals miss even when the code is
correct. A single borderline $z$ is weak evidence either way — re-run with a
different seed before concluding anything.

## 5. A vectorized version

Nothing in a replicate depends on any other replicate, so all `nreps` draws can
be generated at once. `rbinom()` does the work in compiled code:

```{r}
urn_prob_vec <- function(nreps = 100000, b1 = 10, y1 = 8, b2 = 6, y2 = 6) {
  # First draw for every replicate at once: 1 = blue.
  first_blue <- rbinom(nreps, size = 1, prob = b1 / (b1 + y1))

  # Second draw, with urn 2's composition depending on what was transferred.
  p2 <- (b2 + first_blue) / (b2 + y2 + 1)
  mean(rbinom(nreps, size = 1, prob = p2) == 1)
}

urn_prob_vec(nreps = 100000)
```

`prob` is vectorized over the replicates, which is what lets the transfer be
handled without a loop.

## 4. Benchmark

```{r}
#| message: false
bench::mark(
  loop = urn_prob_loop(nreps = 10000),
  vec  = urn_prob_vec(nreps = 10000),
  relative = TRUE,
  check = FALSE
)
```

### Reading the output

`bench::mark()` runs each expression many times and reports thirteen columns,
though only the first nine are printed. From left to right:

| Column | What it is |
|---|---|
| `expression` | The label you gave the expression (`loop`, `vec`), or the code itself if you did not name it. |
| `min` | The fastest single run. Often the most useful summary: it is the run least polluted by garbage collection and by whatever else the machine was doing. |
| `median` | The typical run. Compare it against `min` — a median far above the minimum means the timings are noisy or GC is interfering. |
| `itr/sec` | Iterations per second, i.e. throughput. This is the one column where **bigger is better**; all the timing columns are the opposite. |
| `mem_alloc` | Total memory allocated during one run. Note *allocated*, not peak usage: memory that is allocated and immediately freed still counts, which is exactly what a loop that creates a length-1 vector per iteration does. |
| `gc/sec` | Garbage collections per second. High values mean the expression is generating enough short-lived objects to keep R's collector busy — usually a symptom of allocating inside a loop. |
| `n_itr` | How many times the expression actually ran. `bench` decides this itself from `min_time` (0.5s by default), so a fast expression gets many more iterations than a slow one. |
| `n_gc` | Total garbage collections across all those iterations. |
| `total_time` | Wall-clock time spent on that expression overall — roughly `n_itr` × `median`, and *not* a measure of speed. |

The four hidden list-columns are `result` (the value each expression returned),
`time` (every individual timing), `gc` (the GC events), and `memory` (the full
allocation breakdown). They are there for plotting and debugging:
`ggplot2::autoplot()` on the result uses `time` to show the whole distribution
rather than a single number.

**`relative = TRUE`** rescales `min`, `median`, `itr/sec`, `mem_alloc` and
`gc/sec` so the best performer is `1` and everything else is a multiple of it.
That is what we want here — "the loop is 20× slower" is a portable statement,
whereas "the loop takes 54 ms" is a fact about this laptop. Drop the argument to
see absolute times with their units.

**`check = FALSE`** turns off `bench`'s default insistence that all expressions
return identical values. We need it because the two functions are stochastic and
return different estimates by design. Be careful with it: that check is a free
correctness test, so switch it off only when you know why the results differ,
and satisfy yourself some other way that the fast version is right — here, that
both estimates land within Monte Carlo error of the analytic answer.

**Where did the time go?** Not into the arithmetic — both versions do the same
two comparisons per replicate. The loop pays the R interpreter's per-iteration
overhead 10,000 times over: dispatching `runif`, allocating a length-1 result,
and indexing `blue[i]`. The vectorized version pays that overhead twice, for
whole vectors, and does the per-element work inside C.

**What I would do differently next time:** Return the standard error from the start; without it
there is no way to tell a bug from Monte Carlo noise when the two versions
disagree in the third decimal.

# Part 2: Extending the urn problem (optional)

## 1. Sweep the urn compositions

```{r}
scenarios <- expand.grid(b1 = 5:15, y1 = 8, b2 = 6, y2 = 6)

scenarios$prob <- mapply(
  urn_prob_vec,
  b1 = scenarios$b1, y1 = scenarios$y1, b2 = scenarios$b2, y2 = scenarios$y2,
  MoreArgs = list(nreps = 20000)
)
scenarios$exact <- with(scenarios, urn_prob_exact(b1, y1, b2, y2))

head(scenarios)
```

```{r}
#| label: fig-sweep
#| fig-cap: "Probability the second marble is blue, as urn 1 gets bluer."
plot(scenarios$b1, scenarios$prob,
     type = "b", pch = 19,
     xlab = "Blue marbles in urn 1 (b1)",
     ylab = "Pr(second marble is blue)")
lines(scenarios$b1, scenarios$exact, col = "red", lty = 2)
legend("bottomright", c("simulated", "exact"),
       col = c("black", "red"), lty = c(1, 2), pch = c(19, NA), bty = "n")
```

`mapply()` walks the rows of the grid without a nested loop, and the result stays
a labeled data frame with one row per scenario.

::: {.callout-note collapse="true"}
## Is the functional *faster* than a loop here? (No — and that is worth knowing)

`mapply()` is used here for **readability**, not speed. It is a common myth
that the apply family is the "fast C version" of a `for` loop. It is not.

**The trip-downstairs picture.** R is two layers: the R language you write, and
a fast C layer underneath it. Every call from R down into C costs a fixed
overhead. Think of yourself as a manager with a very fast assistant downstairs,
where each trip down and back takes a minute no matter how small the errand:

- A `for` loop over 100,000 items: you walk downstairs 100,000 times.
- `sapply()` / `Map()` / `lapply()`: the assistant now keeps the checklist and
  the clipboard, so you no longer track which item you are on — but *you* still
  make the decision for each item, so you still walk down 100,000 times. You
  saved the bookkeeping, which was never the expensive part.
- `rbinom(100000, ...)`: you walk down **once**, hand over all 100,000 items,
  and the assistant does the whole batch without you.

The apply family tidies up your bookkeeping; it does not reduce the number of
trips. Real vectorization reduces them to one. Concretely, `lapply()` does the
counting and list-building in C, but the thing being repeated — calling *your*
R function — is still ordinary R, once per element, and that is where nearly
all the time goes.

The sweep above has only 11 scenarios, each running a full simulation, so a
`for` loop and `mapply()` make 11 trips either way and benchmark the same to
within noise. It is the `rbinom()` rewrite in Part 1 that buys the speedup,
because it is the only change that reduces the number of trips:

```{r}
f <- function(x) x + 1
x <- 1:10000

bench::mark(
  loop      = { out <- numeric(length(x))
                for (i in seq_along(x)) out[i] <- f(x[i])
                out },
  vapply    = vapply(x, f, numeric(1)),
  vectorized = x + 1,
  relative  = TRUE, check = FALSE
)[, c("expression", "median", "mem_alloc")]
```

The loop and `vapply()` land within a small factor of each other; the
vectorized version is in a different regime entirely.

**The one loop mistake that does cost you.** Growing the result instead of
preallocating it:

```{r}
grow <- function(n) { out <- c();          for (i in 1:n) out <- c(out, i); out }
prealloc <- function(n) { out <- numeric(n); for (i in 1:n) out[i] <- i;     out }

bench::mark(grow(10000), prealloc(10000), relative = TRUE, check = FALSE)[, c("expression", "median", "mem_alloc")]
```

`grow()` rebuilds the whole vector on every pass — 10,000 progressively larger
copies. This is what gives `for` loops their bad reputation, and it is a
*preallocation* problem, not a loop problem. Switching to `sapply()` happens to
fix it, which is probably how the myth started.
:::

## 2. Generalize to $k$ urns

Represent the urns as a matrix with `blue` and `yellow` columns and one row per
urn, so every index into it reads as a name rather than a number.

```{r}
urn_chain <- function(nreps = 10000, urns = rbind(c(10, 8), c(6, 6), c(4, 9))) {
  urns <- as.matrix(urns)
  colnames(urns) <- c("blue", "yellow")
  rownames(urns) <- paste0("urn", seq_len(nrow(urns)))

  # Replicates are independent, so carry all of them through the chain at once.
  # `transferred` is 1 if the marble moved into the current urn is blue.
  transferred <- rbinom(
    nreps, size = 1,
    prob = urns["urn1", "blue"] / sum(urns["urn1", ])
  )

  for (k in seq_len(nrow(urns))[-1]) {
    n_k    <- sum(urns[k, ]) + 1                    # urn k plus the transfer
    p_blue <- (urns[k, "blue"] + transferred) / n_k
    transferred <- rbinom(nreps, size = 1, prob = p_blue)
  }

  mean(transferred == 1)
}

# Two urns reproduces Part 1.
urn_chain(nreps = 100000, urns = rbind(c(10, 8), c(6, 6)))
truth

# Three urns.
urn_chain(nreps = 100000)
```

**Reading the function.** The physical process is: draw a marble from urn 1,
drop it into urn 2, draw from urn 2, drop it into urn 3, and so on. The only
thing that matters downstream at each step is *the color of the marble that
just moved*, so the whole simulation collapses to a chain of Bernoulli draws and
no individual marble ever needs to be represented.

- The naming step (`colnames()`, `rownames()`) is cosmetic but pays for itself:
  the rest of the body says `urns["urn1", "blue"]` instead of `urns[1, 1]`.
  `as.matrix()` is defensive, in case a data frame is passed.
- The first `rbinom()` draws from urn 1 with probability
  $10/18$. `size = 1` makes it a Bernoulli, and because the first argument is
  `nreps` it returns a *vector* of `nreps` independent draws — all replicates
  are launched at once, which is why there is no outer loop over replicates.
- `seq_len(nrow(urns))[-1]` is $2, \dots, k$, written safely: dropping the first
  element of `seq_len()` still gives an empty sequence when there is only one
  urn, whereas `2:nrow(urns)` would silently run backwards.
- `n_k` is urn $k$'s original count plus the one marble that just arrived — the
  denominator. In `p_blue`, the numerator adds `transferred`, a length-`nreps`
  vector of 0/1, so `p_blue` is itself a *vector* of probabilities, one per
  replicate, differing by whether that replicate's incoming marble was blue.
- `rbinom()` accepts a vectorized `prob` and draws elementwise, so a single call
  advances every replicate one urn. `transferred` is then overwritten with the
  color drawn from urn $k$, ready to feed urn $k+1$.
- After the loop `transferred` holds the color drawn from the *last* urn, so
  `mean(transferred == 1)` estimates $P(\text{final draw is blue})$.

With two urns the loop runs once and reproduces Part 1:
$(10/18)(7/13) + (8/18)(6/13) = 118/234 \approx 0.5043$.

Note that urn $k$'s counts are never decremented after its marble leaves. That
is correct here because each urn is drawn from exactly once; it would need
changing if a marble were ever drawn twice from the same urn.

**Where vectorization stops.** The loop over urns cannot be removed: urn $k$'s
composition depends on the marble drawn from urn $k-1$, so the transfers are
inherently sequential. What *is* still vectorized is the replicate dimension —
each pass through the loop advances all `nreps` chains simultaneously. The loop
therefore runs $k-1$ times, not $k \times \texttt{nreps}$ times, which is the
whole point: vectorize over the independent axis and loop over the dependent one.

## 3. Monte Carlo standard error

```{r}
urn_prob_se <- function(nreps = 100000, b1 = 10, y1 = 8, b2 = 6, y2 = 6) {
  first_blue <- rbinom(nreps, size = 1, prob = b1 / (b1 + y1))
  draws <- rbinom(nreps, size = 1, prob = (b2 + first_blue) / (b2 + y2 + 1))

  phat <- mean(draws)
  c(estimate = phat, se = sqrt(phat * (1 - phat) / nreps))
}

urn_prob_se(nreps = 10000)
```

**How many replicates for SE < 0.001?** Solve
$\sqrt{p(1-p)/n} < 0.001$ for $n$ at $p \approx `r round(truth, 3)`$:

```{r}
p <- truth
n_needed <- ceiling(p * (1 - p) / 0.001^2)
n_needed

urn_prob_se(nreps = n_needed)
```

Roughly a quarter of a million replicates. Note the square-root scaling: cutting
the standard error in half again, to 0.0005, costs four times as many draws. This
is why the vectorized implementation matters — the loop version at this $n$ is
slow enough to be annoying, and at SE = 0.0001 it is unusable.
